---
title: "FTEs"
author: "Brad McNeney"
date: '2017-03-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```


## Science enrollments database

- Read in the science enrollments in wide format.
    - Remember to set your working directory to the 
    `Notes` folder, or a local copy, before 
    reading in the following data.

```{r}
sci <- read.csv("SFUScienceEnrolSummary.csv",stringsAsFactors = FALSE)
head(sci)
```

- The first column has the subject, e.g., `ACMA`, then empty strings
until the next subject, etc. 
- Similarly, there are some course numbers missing, 
when a course had two (or more?) credit values 
over its lifetime (e.g., Stat341 was once a 1-credit course).
- Write a function called `fillEmpty` that takes
an atomic vector as input and returns a filled-in version.

```{r}
fillEmpty<- function(x) {
  for(i in 2:length(x)) {
    if(x[i]=="") {
      x[i] <- x[i-1]
    }
  }
  x
}
sci <- mutate(sci,Subject = fillEmpty(Subject))
sci <- mutate(sci,CrsNum = fillEmpty(CrsNum))
```

- Reshape by gathering the semesterly data into 
rows.
    - Semesters are labelled `X1YYS` where `YY` is the
    year and `S` is the semester, with 1=spring,
    4=summer and 7=fall.

```{r}
scilong <- gather(sci,semester,enrollment,X1074:X1171,na.rm=TRUE)
```

- Extract the semester number from the semester labels.

```{r}
scilong <- mutate(scilong,semester=parse_number(semester))
```

- A full-time equivalent (FTE) is 30 credit hours
of the same subject.

```{r}
scilong <- mutate(scilong,FTEs = CreditHrs*enrollment/30)
```

- Create a `year` variable for the different 
academic years represented in the database.
    - Academic years start in fall and end in summer.
    We only have complete data from 2007/08 through
    2015/16 academic years. 
    - Remove semesters from outside this range.

```{r}
scilong <- filter(scilong,semester >= 1077 & semester <= 1164)
```

- Now create a `year` variable, with value 2008 for 
2007/08, value 2009 for 2008/09, and so on.
    - One could probably come up with something more
    elegant than the following, but it works.

```{r}
scilong <- mutate(scilong,year = 
                    2007*(semester <= 1074)+
                    2008*(semester >= 1077 & semester <= 1084)+
                    2009*(semester >= 1087 & semester <= 1094)+
                    2010*(semester >= 1097 & semester <= 1104)+
                    2011*(semester >= 1107 & semester <= 1114)+
                    2012*(semester >= 1117 & semester <= 1124)+
                    2013*(semester >= 1127 & semester <= 1134)+
                    2014*(semester >= 1137 & semester <= 1144)+
                    2015*(semester >= 1147 & semester <= 1154)+
                    2016*(semester >= 1157 & semester <= 1164)+
                    2017*(semester >= 1167))
```

- Save the `scilong` data frame for future use.

```{r}
save(scilong, file="scilong.RData")
```
