---
title: "Homework 3"
author: "Brad McNeney"
date: '2017-02-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Due date for this homework is Monday March 6, 2017. 

## Shopping data from kaggle

The data for this homework assignment is a small portion
of the data from a kaggle competition called
the Acquire Valued Shoppers Challenge
[https://www.kaggle.com/c/acquire-valued-shoppers-challenge]

The competition was to predict which shoppers will
become repeat buyers of the products they buy 
with coupons. That is, if they buy a product with 
a coupon, what predicts whether they will
become loyal customers of that product?

The data files are described at [https://www.kaggle.com/c/acquire-valued-shoppers-challenge/data]
We will use a subset of these files, available in the Homeworks
folder of SFUStat341 on GitHub:

- `offers.csv`, a CSV file containing information about the 
coupons (a.k.a. offers)
- `trainHistory.csv`, a CSV file with information on coupon use
for customers, including whether they became repeat buyers
- `sample.csv`, a CSV file with 1000 transactions
from the transaction history of
one shopper (the file of all shopper transactions is
22GB -- too large for a homework assignment!)

## Read the CSV files into an SQLite database

The `dbWriteTable()` command we used to create tables
from dataframes can also create tables from files, such
as CSV files.

```{r}
library(DBI);library(RSQLite)
sdb <- dbConnect(RSQLite::SQLite(), "sdb.sqlite")
dbWriteTable(sdb,"offers","offers.csv",overwrite=TRUE)
dbWriteTable(sdb,"history","trainHistory.csv",overwrite=TRUE)
dbWriteTable(sdb,"transactions","sample.csv",overwrite=TRUE)
dbListTables(sdb)
```

**Note** The "overwrite=TRUE" is just 
so that you can re-knit this document without having
to first remove the `sdb.sqlite` file from your computer.

## Questions

Use `DBI` commands to 
do the following questions.

1. (3 marks) Write a batched query to fetch the first
row of the `history` table and save it in a 
one-row data frame called `history1`. 

2. (4 marks) Prior transactions for a customer.
Write a parametrized query, parametrized
by customer ID, that retrieves all transactions for 
that ID from the `transactions` table. 
Use your parametrized query to fetch the transactions
for the ID in `history1` and save in 
a data frame called `oneIDtrans`. Don't forget to 
clear your result set after you fetch it.


3. (3 marks) Prior purchases of products that are 
included in the offers table. Construct a 
query parametrized by customer ID, that inner-joins 
the transactions and offers tables on brand, category
and company, and saves 
the ID, brand, category, company and offer numbers
in a data frame called `priorPurchases`.

    - Note: The brand, category and company columns are 
    in **both** the transactions and offers tables. 
    You can save from, say, the offers table by 
    selecting `offers.brand`, etc. 

Test your query out on the ID in `history1`.
**Do not** clear the result set. 
You will need it for part (4).

4. (3 marks) 
Write a function `priorPurchaseOffer()`
that takes (i) a data frame called
`history1`, like the one you created in part (1) 
and (ii) the parametrized result set from part (3) as
input, creates the `priorPurchases` data frame from 
part (3) for that ID, and returns a logical indicating
whether the offer number in `history1` is in `priorPurchases`.
Test your function out on the `history1`
data frame from part (1). It should return `FALSE`.

