---
title: "Homework 4"
author: "Brad McNeney"
date: '2017-03-09'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started

### Read in the Stat and Act Sci enrolments data 

In lecture 8 we created a data frame
called `stat` that contained data on enrolments
in Stat and Act Sci courses. Copy this to your R working
directory and load it.

```{r, warning=FALSE, message=FALSE}
load("statEnrol.RData")
library(tidyverse)
```


### Course types

1. Stat 1: Stat 101, 201, 203. First course in 
Statistics for non-majors.
2. Stat 2: Stat 302, 305. Second course in Statistics
for non-majors
3. Stat 100: A breadth (B) course in Statistics.
4. Stat 270: Stat 270 is like Stat 1, but is intended for 
Stat and Act Sci majors. It was also a 
required course in Applied Sciences, but 
they are now allowing BUEC 232
5. ACMA: The courses in Actuarial Science
6. Other Stats: These are Stat285 and all the 
upper-division Statistics courses.

## Questions

1. (3 marks) Create a factor variable `courseType` whose value
for each entry of `stat` is one `"Stat1","Stat2","Stat100","Stat270","ACMA","StatOther"`
Add this variable to the `stat` data frame.

2. (2 marks) Group `stat` by `courseType` and `year`
and calculate the 
sum of the `FTEs` for each group. Save
your summary in a data frame called `statByType`

3. (3 marks) Use `ggplot()` to draw a scatterplot
of `FTEs` by `year`, stratified on `courseTypes`. Include
scatterplot smoothers, but no standard errors. Add 
an informative title to your plot.

4. (2 marks) Extract from `stat` the courses in the 
`StatOther` category. These courses are Stat 285
(Stat 2 for Stats majors), and the upper-division
courses for Stats majors and minors. (One exception is
Stat 403 which is more of a service course than a course
for majors and minors, but its enrolments are relatively small
and I've left it in with the courses for majors and minors.)
Experiment with different
graphical displays that might shed light on the changes 
in enrolments over time. **There is no right or wrong "answer" here**, just experiment.

FYI, the courses that are considered accessible to minors are
340/341/342, 403, 445, 475 and 485.
With the exception of Stat 403, these were 
introduced in the 2012/13 academic year, _after_ the
increase in enrolments in the `statOther` category.


5. (3 marks) Repeat the plot in question (3) but instead of plotting 
the _number_ of FTEs
for a given year and course type, plot FTEs as a proportion
of all FTEs for each year. That is, the point for 
Stat1 in 2008 should be the proportion of all Stat and Act Sci
FTEs that were due to Stat1 courses.
of FTEs for each year
plot the proportion of all FTEs for for 
each course type with the proportion of FTEs out of all
FTEs in Stat and Act Sci.
