---
title: "Homework 5"
author: "Brad McNeney"
date: '2017-03-24'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(stringr)
library(nycflights13)
library(DBI)
```

## Questions

1. (1 mark) Use the `nycflights13_sqlite()` function to copy the `nycflights13`
data to an SQLite database. Record the location of the database file
on your computer. You may want to use the `path` 
argument of `nycflights13_sqlite()` to specify a file path 
that is easy to remember.

2. (1 mark) Use `DBI` functions to establish a connection to the database from 
part (1) and list its tables. 


3. (2 marks) The `flights` and `airlines` tables are the same as the 
dataframes of the same name in the
`nycflights13` package. See the R help file for `flights` and `airlines`
for information on their columns
Use `DBI` to perform an inner join between the `airlines` table
and the `flights` table on `carrier`. In your SQL command 
select only the `carrier`, `name`, `month`, `day`, `dep_time`
and `dep_delay` variables.
Save your dataframe in `flights`.


4. (2 marks) Remove Inc., Co. and Corporation from the airline names in the
`name` variable of `flights`. Then shorten the phrases "Airlines",
"Air Lines", and "Airways" to just "Air". Print out the 
unique values of `flights$name`.

5. (2 marks) Select the flights from January. Calculate the median departure 
delay for each airline on each day and save the results in a 
dataframe called fldelaysJan. Plot median delay versus day, 
with different facets for each airline. Use the `name` variable
to specify airlines (instead of `carrier`) so that the labels
on your facets are informative.

6. (2 marks) The plot in part (4) is not very informative because one airline 
had a very long median flight delay on one day in January. Remove this 
one observation from `fldelaysJan`. Also remove the airline
that had flights on only one day. Now redo the plot in (4).

7. (3 marks) Identify the four airlines with the greatest median
delays. Plot these four as facets on a single plot of median delay 
versus day, and plot the remaining airlines as facets on a separate plot of 
median delay versus day.
Add informative titles to your two plots.

8. (2 marks) Select the flights from January 1 and do line plots
of departure delay by departure time, with different airlines
in different facets.

9. (2 marks) Redo the plot in part (8) for the airlines that had
at least 100 flights on Jan 1.
