---
title: "Homework 3"
author: "Homework Group X: Name1, Name2, Name3"
date: '2017-03-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started

### Read in the Stat and Act Sci enrolments data 

In lecture 7 we created a data frame
called `stat` that contained data on enrolments
in Stat and Act Sci courses and saved it in a .RData file 
called `statEnrol.RData`. Copy this .RData file to your
current R working
directory and load it.

```{r, warning=FALSE, message=FALSE}
load("statEnrol.RData")
library(tidyverse)
```


### Course types

1. Stat 1: Stat 101, 201, 203. First course in 
Statistics for non-majors.
2. Stat 2: Stat 302, 305. Second course in Statistics
for non-majors
3. Stat 100: A breadth (B) course in Statistics.
4. Stat 270: Stat 270 is like Stat 1, but is intended for 
Stat and Act Sci majors. It was also a 
required course in Applied Sciences, but 
they are now allowing BUEC 232
5. ACMA: The courses in Actuarial Science
6. Other Stats: These are Stat285 and all the 
upper-division Statistics courses.

## Questions

1. (3 marks) Create a factor variable `courseType` whose value
for each entry of `stat` is one `"Stat1","Stat2","Stat100","Stat270","ACMA","StatOther"`
Add this variable to the `stat` data frame.

2. (2 marks) Group `stat` by `courseType` and `year`
and calculate the 
sum of the `FTEs` for each group. Save
your summary in a data frame called `statByType`

3. (1 mark) Use the `split()` function from base R to 
split `statByType` on `courseType` and save the results in 
a list called `sp.statByType`.

4. (2 marks) Create a function `regFY()` that 
takes a data frame  as an argument and uses `lm()`
to fit a least squares
regression of `FTEs` on `year`. You may assume that 
the data frame argument includes columns `FTEs` and 
`year`. Also create a function `estSlope()` that 
extracts the coefficient of `year` from the output 
of `lm()`.


5. (2 marks) Use `map()` from the `purrr` package
and your `regFY()` function to fit a least squares 
regression of `FTEs` on `year` for
each element of `sp.statByType` and then
use `map_dbl()` and `estSlope()` to extract the regression summaries
for each fitted model.
