---
title: "Homework 4"
author: "Homework Group X: Name1, Name2, Name3"
date: '2017-03-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(stringr)
library(nycflights13)
```

## Questions

0. (0 marks) If you haven't already, install
the `nycflights13` package and read the 
help files on the 
`flights` and `airlines` tables.

1. (1 mark) 
Use `dplyr` functions to perform an inner join between the `airlines` table
and the `flights` table on `carrier` and 
then select the variables
`carrier`, `name`, `month`, `day`, `dep_time`
and `dep_delay`.
Save your dataframe in `flights`.

2. (2 marks) Remove Inc., Co. and Corporation from the airline names in the
`name` variable of `flights`. Then shorten the phrases "Airlines",
"Air Lines", and "Airways" to just "Air". Print out the 
unique values of `flights$name`.

3. (2 marks) Select the flights from January. Calculate the median departure 
delay for each airline on each day and save the results in a 
dataframe called fldelaysJan. Plot median delay versus day, 
with different facets for each airline. Use the `name` variable
to specify airlines (instead of `carrier`) so that the labels
on your facets are informative. Do **not** free
the scales on the facets.

4. (2 marks) The plot in part (4) is not very informative because one airline 
had a very long median flight delay on one day in January. Remove this 
one observation from `fldelaysJan`. Also remove the airline
that had flights on only one day. Now redo the plot in (4).
Again, do **not** free the scales on the facets.

5. (3 marks) Identify the four airlines with the greatest median
delays. Plot these four as facets on a single plot of median delay 
versus day, and plot the remaining airlines as facets on a separate plot of 
median delay versus day.
Add informative titles to your two plots.

