---
title: "Week 6 exercises"
author: "Brad McNeney"
date: '2018-02-08'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Set up DB 

```{r}
wdb <- dbConnect(RSQLite::SQLite(), "wdb.sqlite")
STATION <- data.frame(ID=c(13,44,66),
    City = c("Phoenix","Denver","Caribou"),
    State = c("AZ","CO","ME"),
    Lat_N = c(33,40,47),
    Long_W = c(112,105,68))
STATS <- data.frame(row = 1:6,
    ID = c(13,13,44,44,66,66),
    Month = c(1,7,1,7,1,7),
    Temp_F = c(57.4,91.7,27.3,74.8,6.7,65.8),
    Rain_I = c(0.31,5.15,0.18,2.11,2.1,4.52))
dbWriteTable(wdb,name='STATION', value = STATION, overwrite=TRUE) 
dbWriteTable(wdb,name='STATS', value = STATS, overwrite=TRUE)
dbListTables(wdb)
```

### Exercises

Using the database connection `wdb` do the following exercises.

1. Select the stations from the state of Colorado
2. Add Vancouver as 
a new station in the `STATION` table
(Use BC as the "State", latitude 49 and
longitude 123)
3. Do an inner join that returns a table with city,
state, and temperatures for July from 
cities at latitude 40 or more.
4. Repeat (3) as a left join. 
5. Do a left join that returns a table with city,
state, and temperatures for July from 
cities at north latitude 40 or more.
(Vancouver excluded because
we filter on month, and month is missing for Vancouver.)
6. Do a left join that returns a table with city,
state, and temperatures from 
cities at north latitude 40 or less.


