---
title: "Week 1 Exercises"
author: "Brad McNeney"
date: '2019-01-04'
output: 
  html_document:
    fig_width: 6
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory analysis of gapminder data

Load the gapminder data. We will also need
an R package called `dplyr` that will facilitate exploratory analyses and `ggplot2` to allow visualization.

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)
```
Take a look at the top and bottom few
lines of raw data.

```{r}
head(gapminder)
tail(gapminder)
summary(gapminder)
```

Type `help("gapminder")` in the R console for information
about the gapmider dataset.

We will explore the life expectancy variable for 
the year 2007. First filter the data to just 2007.

```{r}
gapminder07 <- filter(gapminder, year == 2007)
head(gapminder07)
```

In R, the `<-` is the assignment operator that creates 
new variables/datasets.

## Life expectancy by continent
Calculate median life expectancy, first overall, 
and then by continent.

```{r}
summarize(gapminder07, median(lifeExp))
by_cont <- group_by(gapminder07, continent)
summarise(by_cont, median(lifeExp))
```

In the above commands, `group_by()` creates a new data set
with observations grouped by continent.

We can visualize the median life expectancies.

```{r}
medL <- summarize(by_cont, median(lifeExp))
plot(medL)
```

What is "Oceania"?

```{r}
filter(gapminder07,continent == "Oceania")
```


The `dplyr` package allows for us to "chain" the filter,
grouping and summary commands. The following is 
an equivalent way to construct `medL`:

```{r}
medL <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarise(medLifeExp = median(lifeExp))
```

## Life expectancy over time

First look at African countries
```{r}
medLA <- gapminder %>%
  filter(continent == "Africa") %>%
  group_by(country) %>%
  summarise(medLifeExp = median(lifeExp))
```

Look at a subset of countries with the lowest and highest median life expectancies.
```{r}

filter(medLA,medLifeExp<40)
filter(medLA,medLifeExp>60)
cc = c("Angola","Guinea-Bissau","Sierra Leone",
       "Mauritius","Reunion","Tunisia",
       "Mexico") # Mexico for comparison
```

Plot life expectancy over time. 
Illustrate chaining of filtering (on country) and ggplot.

```{r}
gapminder %>%
  filter(country %in% cc) %>%
  ggplot(aes(x=year,y=lifeExp,color=country)) +
   geom_point() +
   geom_smooth(method = "lm")
```

Here's another interesting plot of life expectancy over time:

```{r}
gapminder %>%
  filter(continent == "Oceania") %>%
  ggplot(aes(x=year,y=lifeExp,color=country)) +
  geom_point() +
  geom_smooth(method = "loess", span=3/4)
```
What happend in the mid-1970s in Australia?

## Life expectancy versus per capita GDP

First try a simple scatterplot of `lifeExp` versus 
`gdpPercap`.

```{r}
qplot(gdpPercap,lifeExp,data=gapminder07)
```

It is hard to make sense of the pattern in `lifeExp` versus `gdpPercap`.
Try grouping the data by continent. (Note: This does not 
use our `by_cont` data set. We'll talk about why later.)

```{r}
qplot(gdpPercap,lifeExp,data=gapminder07,color = continent)
```

Add regression lines for each continent. Doing so uses a
more complicated graphing function from `ggplot2`.

```{r}
ggplot(gapminder07, aes(x=gdpPercap,y=lifeExp,color=continent)) +
 geom_point() +
 geom_smooth(method = "lm", se=FALSE)
```

