---
title: "Week 5 exercises"
author: "Brad McNeney"
date: '2017-02-06'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Background for exercises

Suppose you have a directory that contains thousands of 
files from a genetics experiment. The file names are
of the form `chrNgeneC.txt`, where `chr` is short
for chromosome, `N` can be a number
between 1 and 22, or the letters `X` or `Y` and 
`C` is a character string of arbitrary length. Here is a sample of 
what these filenames might look like.
```{r}
fl <- c("chr1geneC1orf49.txt","chr1geneAMPD2.txt",
        "chr2geneCOL4A4.txt","chr2geneHADHB.txt",
        "chrXgeneCXorf49.txt")
print(fl)
```

### Exercise 1: Function to find the files from a chromosome

Write a function called `findChrFiles()`
that takes the following inputs: (i) a character 
vector (such as the example `fl` above) 
with the filenames, and (ii) a 
character string in the set 
`cnum <- c(as.character(1:22),"X","Y")`
and returns all filenames that include that chromosome.  Have
your function check that the chromosome identifier is 
in `cnum` and return the error message "Invalid chromosome" if not. 

### Exercise 2: Extract gene names

Write a function `findGeneNames()` that takes a list of filenames and 
returns the gene names found in the filenames. 
Hints: You can use regular expressions and
`str_extract()` to extract strings of the 
form `geneC.txt`. Then how do you trim off the leading 
`gene` and the trailing `.txt`?


### Exercise 3: read in files

Write a function `readORF()`
that takes a chromosome identifier in 
the set `cnum` defined in exercise 1 and a vector of 
gene names, and reads in all the files from that
chromosome with a gene name that contains the string `orf`.
Assume that the files all contain tabular data that is
read in correctly by `read.table()` with its default settings.
Save the data from these files in a list that has one 
list element per gene, and with list elements named
by gene. For example, "chrXgeneCXorf49.txt" should
get read into a list element with name "CXorf49".
Don't bother with error checking the inputs.
