---
title: "Week 8 exercises"
author: "Brad McNeney"
date: '2018-03-07'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## HIV prevalence from WHO

- Estimated HIV prevalence was obtained from the 
`gapminder` website https://www.gapminder.org/data/
    * Estimated number of people living with HIV per 100 population of age group 15-49.
    * Original data source is the UNAIDS online database at 
    http://www.aidsinfoonline.org

```{r}
library(tidyverse)
hiv <- read.csv("HIVprev.csv",stringsAsFactors = FALSE)
```

1. The first column of the data frame is the country, but it
has been named`Estimated.HIV.Prevalence.....Ages.15.49.1`.
Use the `rename()` function to rename this column  `Country`.

2. The data from 1979 to 1989 is very sparse. Remove these
columns from the data frame.

3. Gather the yearly prevalence estimates into key, value pairs
with `year` as the key and `prevalence` as the value. Save
the result as `hivlong`.

4. Sort `hivlong` on `Country`.

5. Extract the years from the `year` column and 
coerce to numeric.

6. Use the `na.omit()` function to remove rows of `hivlong` 
with missing prevalence and filter to rows
that contain data from the years 1990 to 2000, inclusive.


7. Write a function `hivregr()` that takes data for a single
country, fits a least squares regression of `prevalence` on `year`
and returns the estimated regression slope.


8. Split the `hivlong` data frame on `Country` and
use `map_dbl()` to apply `hivregr()` to the data from 
each country. Save the results in a vector `slopes`.
Print the countries with estimated slope greater than one.

9. Repeat part, but instead of defining a function `hivregr()`
outside `map_dbl()`, define the function within the call to 
`map_dbl()` and use all available short-cuts.