---
title: "Week 6 exercises"
author: "Brad McNeney"
date: '2018-02-08'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Set up DB 

```{r}
library(DBI)
wdb <- dbConnect(RSQLite::SQLite(), "wdb.sqlite")
STATION <- data.frame(ID=c(13,44,66),
    City = c("Phoenix","Denver","Caribou"),
    State = c("AZ","CO","ME"),
    Lat_N = c(33,40,47),
    Long_W = c(112,105,68))
STATS <- data.frame(row = 1:6,
    ID = c(13,13,44,44,66,66),
    Month = c(1,7,1,7,1,7),
    Temp_F = c(57.4,91.7,27.3,74.8,6.7,65.8),
    Rain_I = c(0.31,5.15,0.18,2.11,2.1,4.52))
dbWriteTable(wdb,name='STATION', value = STATION, overwrite=TRUE) 
dbWriteTable(wdb,name='STATS', value = STATS, overwrite=TRUE)
dbListTables(wdb)
```

### SQL Exercises

Using the database connection `wdb` do the following exercises.

1. Select the stations from the state of Colorado


2. Add Vancouver as 
a new station in the `STATION` table
(Use BC as the "State", latitude 49 and
longitude 123)


3. Do an inner join that returns a table with city,
state, month and temperatures 
from cities at latitude 40 or more.

4. Repeat (3) as a left join. 


5. Do a left join that returns a table with city,
state, month and temperatures for July from 
cities at north latitude 40 or more.
(Vancouver excluded because
we filter on month, and month is missing for Vancouver.)


### Exercises with `dplyr`

Using the `STATION` and `STATS` data frames defined above, 
answer the following.

1. Filter `STATION` to just the stations from the state of Colorado 


2. Select the City and State from the stations
using the `select()` command from `dplyr`:


3. Add Vancouver as 
a new station in the `STATION` data frame.
Use BC as the "State" and the latitude 49 and
longitude 123. 

4. Do an inner join that returns a table with city,
state, month and temperatures for July from 
cities at north latitude 40 or more.

5. Repeat (3) as a left join.



