---
title: "Week 9 exercise"
author: "Brad McNeney"
date: '2017-03-04'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data on SFU Stat and Act Sci students

- The following data are from a summary of 
Stat and Act Sci students made available
by SFU's Institutional Research and Planning office
at [https://www.sfu.ca/content/dam/sfu/irp/departments/stat_tables.pdf].

```{r, warning=FALSE, message=FALSE}
stat <- data.frame(
  year = 12:16,
  FTEs = c(446.47,484.8,483.53,443.97,466),
  majors = c(213,245,260,228,233),
  minors = c(17,27,62,77,111),
  p.intl = c(.4,.42,.51,.53,.54)) 
head(stat)
```

- `year` is the end-year of an academic year; e.g., 
12 = 2011/2012, and so on up to 16 = 2015/2016.
- `FTEs` stands for full-time equivalent, which is 30 credit
hours of classes in statistics and actuarial science.
    - For example, if there are 120 students in Stat 341,
    each receiving 2 credit hours, then this amounts to 
    $120\times 2/30 = 8$ FTEs.
- `majors` is the number of declared Stat or Act Sci majors.
- `minors` is the number of declared Stat or Act Sci minors. 
- `p.intl` is the proportion of interantional students.

1. Load the packages in the `tidyverse` and 
reshape the `FTEs`, `majors` and `minors`
columns into a key-value pair of columns
with the key variable labeled as 
`type` and value variable labeled as `students`.

2. Plot `students` versus `year` by group `type`,
with different colors for each type of student.
Add a scatterplot smoother **without** standard
error bands.
Interpret the graph. 

3. Add the following variables to the `stat` data frame:

- `total = majors+minors`
- `international = total*p.intl`
- `domestic = total*(1-p.intl)`

4. Reshape the `total`, `international` and `domestic`
columns into key-value pairs with key labeled
`type` and value labeled `students`. 

5. Plot  `students` versus `year` by group `type`,
with different colors for each type of student.
Add a scatterplot smoother **without** standard
error bands.
Interpret the graph.