---
title: "Week 3 exercises"
author: "Brad McNeney"
date: '2019-01-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Control flow

1. Rewrite the following function so that 
it uses a `while()` loop instead of the 
`for()` loop and `break` statement. Your while-approach
will not require the `maxit` upper limit on the 
number of iterations.
```{r}
rtruncNormal <- function(thresh = 2, maxit=1000) {
  x<-NULL
  for(i in 1:maxit) {
    xnew <- rnorm(n=1)
    if(xnew>thresh) {
      break
    }
    x <- c(x,xnew)
  }
  x
}
set.seed(1234)
rtruncNormal()
```


## Dates

2. Write an R command to read a CSV file called
`trainHistory.csv` data into an 
R data frame called `thdat` and print the first 6 rows.
You can assume that the CSV file is in your
working directory.


3. Suppose the last column of `thdat`
is called `offerdate` and that
it represents dates in the form yyyy-mm-dd. Give
an R command to replace the `offerdate` column in 
`thdat` with an object of class `Date`.

## Character strings

4. Consider the following vector of character 
strings:
```{r}
gg <- c("TT","TC","TT","CT","TT","TT")
```
Write a code chunk that (i) extracts the first
letter and saves it in a vector `l1`,
(ii) extracts the second allele and saves it 
in a vector `l2`, (iii) counts the number of instances
of each letter in `l1` and `l2` combined.

