---
title: "Week 8 exercises"
author: "Brad McNeney"
date: '2017-02-26'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## SQL-like operations in R

- These are the exercises from the `lec06.Rmd` document, 
with a few extras added. You will need to load the `dplyr` package.

```{r}
library(dplyr)
```


- We will work with the `STATION` and `STATS` tables that
we used in the SQL lecture.
    - Because we will be working in R, we need to take
    care about coercing strings to factors.

```{r}
STATION <- data.frame(ID=c(13,44,66),
    City = c("Phoenix","Denver","Caribou"),
    State = c("AZ","CO","ME"),
    Lat_N = c(33,40,47),
    Long_W = c(112,105,68),stringsAsFactors=FALSE)
STATS <- data.frame(row = 1:6,
    ID = c(13,13,44,44,66,66),
    Month = c(1,7,1,7,1,7),
    Temp_F = c(57.4,91.7,27.3,74.8,6.7,65.8),
    Rain_I = c(0.31,5.15,0.18,2.11,2.1,4.52))
```

1. Use the `filter()` function to
select the stations from the state of Colorado

2. Select the City and State from the stations
using the `select()` command from `dplyr`:

3. Add Vancouver as 
a new station in the `STATION` data frame.
Use BC as the "State" and the latitude 49 and
longitude 123. 

4. Use `merge()` and whatever else you need to 
do an inner join that returns a table with city,
state, and temperatures for July from 
cities at north latitude 40 or more.

5. Repeat (4) with `inner_join()` instead of `merge()`.

6. Repeat (4) as a left join with `left_join()`.
