---
title: 'Lecture 9'
author: "Brad McNeney"
date: '2018-03-15'
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


## Load packages

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
data(diamonds)
library(dplyr)
library(gapminder)
data(gapminder)
gapminder <- mutate(gapminder, 
                    log10Pop = log10(pop), 
                    log10GdpPercap = log10(gdpPercap))
```

## References

- ggplot2 cheatsheet at [https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf]
- Wickham (2009) ggplot2: Elegant graphics for data analysis, Chapters 4 and 5.
- Chang (2012) R graphics cookbook. Available at  [http://www.cookbook-r.com/Graphs/]

## More details on `ggplot2`

\small

- **Layers** 
    - data, aesthetic mapping, geom, statistical transformation 
    and position adjustment (to be defined)
- Tools for working with layers
- Scales, axes and legends
- Faceting 

## Basic plot for demonstrations

\scriptsize

```{r, fig.width=4,fig.height=2}
p <- ggplot(diamonds,aes(x=carat,y=price,colour=cut)) +
  geom_point()
names(p)
p
```


## Data

\small

- The data must be a data frame
- A **copy** of the data is stored in the plot
object (changes to the source dataframe do not change plot)
- Possible to change the data of a plot
object with `%+%`, as in

\scriptsize

```{r,fig.width=4,fig.height=2}
set.seed(123)
subdiamonds <- sample_n(diamonds,size=100)
p <- p %+% subdiamonds
p
```

## Different data in different layers

\small

- Can specify data for a layer to use.

\scriptsize

```{r,fig.width=4,fig.height=2}
p + geom_smooth(data=diamonds, method="lm")
```


## Aesthetic mappings

\small

- A description of how variables will be perceived
on a plot.
    - E.G., `aes(x=carat,y=price,color=cut)`
- Can specify default aesthetics in 
the initialization, or specific aesthetics
in the layers
    - When specified in the layers they over-ride
    the defaults
    
\scriptsize

```{r, fig.width=4, fig.height=2}
p<-p + geom_point(aes(color=clarity))   
```


## Aesthetic over-ride

\small

- Overriding affects only 
the layer, not the default scales
    - Unexpected behaviour: titles on legends and axes are
    taken from the default scale **and** layer; e.g., cut and
    clarity categories
 
\scriptsize

```{r, fig.width=4, fig.height=2.7}
p
```

## Aesthetic over-ride, cont.

\small

- To modify the scale,
set with a `scale` function (more later).

\scriptsize

```{r, fig.width=4, fig.height=2.7}
clarityscale <- c("SI2","SI1","VS1","VS2","VVS2","VVS1","I1","IF")
p + scale_color_discrete(name="clarity",breaks=clarityscale)
# or p + scale_color_hue(name="clarity",breaks=clarityscale)
```

## Setting vs mapping 

\small

- An alternative to mapping aesthetics to 
variables is to set the aesthetic to a constant.
    - Set with the layer _parameter_, rather 
    than mapping with `aes()`
    - E.G., set the color of points 
    on a plot to dark blue

\scriptsize

```{r, fig.width=4, fig.height=2.5}
ggplot(subdiamonds,aes(x=carat,y=price)) + geom_point(color="darkblue")
# Compare to using geom_point(aes(color="darkblue"))
```

## Setting vs mapping, cont. 

\small

- Don't use `aes()` if you want a paramter.
    - e.g., `aes(color="darkblue")` creates a variable whose only
    value is "darkblue"

\scriptsize

```{r, fig.width=4, fig.height=2.5}
ggplot(subdiamonds,aes(x=carat,y=price)) + 
 geom_point(aes(color="darkblue"))
```


## Grouping

\small

- Many `geoms` in ggplot2 group observations
(rows of the dataframe).
    - E.G., a boxplot of a quantitative 
    variable by a categorical variable groups
    observations by the categorical variable. 
- Default `group` is combinations (interaction)
of all categorical variables in the aesthetic specification.
- If this is not what you want, or if there
are no categorical variables, specify `group` yourself.

## Grouping to plot time series

\small

- For plotting time series (multiple measurements
on each observational unit) we want to group
by observational unit.

\scriptsize

```{r, fig.width=4, fig.height=2.5}
ggplot(gapminder,aes(x=year,y=lifeExp,group=country)) + geom_line(alpha=.5)
```

## Different groups on different layers

\small

- To add summaries by continent, we need
to specify different groups on different layers.

\scriptsize

```{r, fig.width=4, fig.height=2.2}
ggplot(gapminder,aes(x=year,y=lifeExp,group=country,
                     color=continent)) + geom_line(alpha=0.2) + 
  geom_smooth(aes(x=year,y=lifeExp,group=continent),se=FALSE)
```

## Using `interaction()` to specify groups

\small

- Could do boxplots of life expectancy by year 
**and** continent. (Not recommended -- too busy --
just using for illustration.)

\scriptsize

```{r, fig.width=4, fig.height=2.2}
ggplot(gapminder,aes(x=year,y=lifeExp, 
         group=interaction(year,continent),color=continent)) + 
 geom_boxplot()
```

## Overriding `group` on a layer

\small

- Boxplots of life expectancy by year 
and continent on one layer, by 
continent alone on another. 

\scriptsize

```{r, fig.width=4, fig.height=2.2}
ggplot(gapminder,aes(x=year,y=lifeExp, group=interaction(year,continent),
                     color=continent)) + geom_boxplot() +
  geom_smooth(aes(group=continent), se=FALSE)
```

## Geoms

\small

- These are the shapes you want on the plot.
    - See the list of geoms on the cheatsheet.
- Each has a set of aesthetics that are required
for drawing and a set of aesthetics that 
it understands.
    - E.G., `geom_point()` requires x and y position, 
    and understands color, size and shape.
- Aesthetics can also be passed to geoms as parameters.
    - Recall difference between `geom_point(color="darkblue")` and `geom_point(aes(color="darkblue"))`
- Each geom also has a default statistic (stat)
and positional adjustment.
    - More on these next.

## Stats

\small

- Stats are statistical summaries of groups
of data points. E.G.,
    - `stat_smoother()` is a moving average
    of the y positions over windows of x positions.
    - `stat_bin()` is a binning of a quantitative 
    variable into bins of a given width
    - See the cheatsheet for a list.
- Stats create new variables, and these
variables can be mapped to aesthetics.
    - E.G., `stat_bin()` creates
    the variables `count`, `density` and `x`
    - Enclose derived variable name with `..` 
    to use. 

\scriptsize

```{r, fig.width=4, fig.height=2, warning=FALSE}
p <- ggplot(gapminder,aes(x=lifeExp)) + 
  geom_histogram(aes(y= ..density..))
```


## Position adjustment

\small

- See cheatsheet for list.
- Default for most geoms is no adjustment ("identity")
- Adjustment to x and/or y position, such 
as "jitter" can reduce overplotting.
- Boxplots in recent plot of gapminder data
were "dodge"d. 
- Histograms of a continuous variable 
by a categorical grouping variable are
"stack"ed by default.

\scriptsize

```{r,fig.width=4, fig.height=1.5, message=FALSE}
gdat <- filter(gapminder,year==1952)
ggplot(gdat,aes(x=lifeExp,color=continent)) + geom_histogram()
```

## More details on `ggplot2`

\small

- Layers 
- **Tools for working with layers**
- Scales, axes and legends
- Faceting 

## Tools for working with layers

- Many possible topics
- In the interest of time, focus on a few 
    - displaying distributions
    - adding error bars and other measures of 
    uncertainty
    - annotating a plot

## Displaying distributions

- The standard histogram is `geom_histogram()`.
    - Displays counts by default, but we have
    seen how to display as density
    - Density is better for comparing the
    shape of distributions
- To include group information, can stack bars
(see previous example),
use faceting to produce separate histograms,
or superpose density histograms.

## Histograms with faceting

\scriptsize
```{r, fig.width=4, fig.height=3}
gdat <- filter(gdat,continent != "Oceania")
h <- ggplot(gdat,aes(x=lifeExp)) 
h + geom_histogram(aes(y= ..density..), binwidth=5) + facet_grid(continent ~ .)
```

## Histograms superposed

\scriptsize


```{r, fig.width=4, fig.height=3}
h + geom_freqpoly(aes(y=..density..,color=continent), binwidth=5)
```

## Density estimation

\small

- `geom_density()` plots a density estimate
    - Think of adding up small normal densities
    centred at each observed datapoint, with 
    the width of each distribution a user-defined
    parameter
- Can superpose multiple density plots.

\scriptsize

```{r, fig.width=4, fig.height=2.2}
h + geom_density(aes(color=continent))
```

## Violin plots

\small

- Instead of superposing, dodge density 
estimates with a violin plot.
    - Violins are density estimate and
    its mirror image displayed vertically

\scriptsize

```{r, fig.width=4, fig.height=2.2}
h + geom_violin(aes(x=continent,y=lifeExp)) 
```

## Adding measures of uncertainty

\small

- `geom_smooth()` includes pointwise error bands by default.
- For factors can add error bars.

\scriptsize

```{r,fig.width=4,fig.height=2,warning=FALSE}
gfit <- lm(lifeExp ~ continent,data=gdat)
newdat <- data.frame(continent=c("Africa","Americas","Asia","Europe")) 
mm <- data.frame(newdat,predict(gfit,newdata=newdat,interval="confidence"))
ggplot(mm,aes(x=continent,y=fit)) +
  geom_point() +  geom_errorbar(aes(ymin=lwr,ymax=upr))
```

## Measures of uncertainty with stat summaries

\small

- Variety of built-in summaries, or can write your own (not
covered).
- Summarize y for different values of x or bins of x values.

\scriptsize

```{r, fig.width=4, fig.height=2.2}
ggplot(gdat,aes(x=continent,y=lifeExp)) + 
 geom_violin() + # superpose over violin plot
 stat_summary(fun.data="mean_cl_normal",color="red") 
```

## Annotating a plot

\small

- Basic tools for annotating are
    - `geom_text()` and `geom_label()` to add text
    - Geoms such as `geom_abline()` to add line segments (see cheetsheet)
    - `labs()` for adding axis labels,
    titles, and captions
    - `annotate()` to add annotations using aesthetics
    passed as vectors to the function, rather than mapped from a dataframe.
- Can add annotations one at a time or many at a time
    - to add many at a time, create a data frame

## Many annotations

\scriptsize

```{r, fig.width=4,fig.height=2.5}
gm07 <- filter(gapminder, year ==2007)
topOilbyGDP <- c("Kuwait","Guinea","Norway","Saudi Arabia")
gdpOil <- filter(gm07,country %in% topOilbyGDP)
ggplot(gm07,aes(x=gdpPercap,y=lifeExp)) + geom_point() +
  geom_text(data=gdpOil,aes(label=country))
```


## More details on `ggplot2`

- Layers 
- Tools
- **Scales, axes and legends**
- Faceting

## Scales

\small

- Scales are mappings from the data to the graphics device
    - domain of `continent` is the five continents, range is
    the hexidecimal of the 
    five colors represented on the graph
    - domain of `lifeExp` is 23.599 to 82.603, range is [0,1],
    which `grid` converts to a 
    range of vertical pixels on the graph.
    - legends and axes provide the inverse mapping

\scriptsize

```{r, fig.width=4, fig.height=1.7}
p <- ggplot(gapminder,aes(x=gdpPercap,y=lifeExp,color=continent)) +
  geom_point(alpha=0.5) + coord_trans(x="log10")
p
```

## Steps to map domain to range

\small

- Transformation: For continuous domains, we may want 
a transformation of the data
    - E.G., log base 10 of GDP, specified by `coord_trans(x="log10")`
- Training: If multiple layers will be plotted, 
take the union of the domains of each
    - Can over-ride training and specify manually (next slide)
- Mapping: Apply the scaling function to map data
values to what we see.


## Axis labels and limits

\small

- `breaks` are the locations of the tick marks on the 
axes
- `limits` are the limits of the axes

\scriptsize

```{r,fig.width=4,fig.height=2}
p + scale_x_continuous(breaks=c(1000,10000,100000)) +
 scale_y_continuous(limits = c(0,100))
```

## Types of scales

\small

- In the previous example we saw position scales 
that control x and y positions of objects
- Other scales control colour and fill, shape, 
line type and size.
    - See the cheetsheet for a list

## Color scale settings and legends

\small

- For discrete variables used to set the colour
aesthetic, colour scale settings affect the
legend labels and titles 

\scriptsize
```{r, fig.width=4,fig.height=2.5}
p + scale_color_discrete(name="Continent",
  labels=c("Africa","North and South America",
           "Asia and Middle East","Europe","Australia and New Zealand"))
```


## More details on `ggplot2`

- Layers 
- Tools 
- Scales
- **Faceting**
    
## Faceting

\small

- Facets are panels that show plots for subsets of the
data defined by groups.
- Layout of facets can be as a grid or a ribbon.
- Use a subset of the `mpg` dataset from `ggplot2` 
to illustrate

\scriptsize
```{r}
data(mpg)
mpgsub <- subset(mpg,cyl !=5 & drv %in% c("4","f"))
```


## Facet grid

\small

- Lay out plots in a 2d grid, specified by a 
formula.

\scriptsize

```{r, fig.width=4,fig.height=2.5}
p <- ggplot(mpgsub,aes(x=cty,y=hwy)) +
  geom_point() + facet_grid(drv~ cyl)
p
```

## Freeing the scales across rows/columns of panels

\scriptsize

```{r, fig.width=4,fig.height=3}
p <- ggplot(mpgsub,aes(x=cty,y=hwy)) +
  geom_point() + facet_grid(drv~ cyl, scales="free")
p
```

## Facet wrap

\scriptsize

```{r, fig.width=4,fig.height=3}
ggplot(gapminder,aes(x=log10GdpPercap,y=lifeExp,color=continent)) + 
  geom_point() + facet_wrap(~year,ncol=4)
```

