---
title: "lecture 5"
author: "Brad McNeney"
date: '2017-01-29'
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Working with character strings

\small

Fixed, or literal strings:

- count the number of characters in a string
- detect (yes/no) or find (starting position) substrings 
- extract and substitute substrings
- split and combine strings

String patterns:

- detect, find, extract and substitute

Througout, illustrate "base" R utilities and those
from the `stringr` package

## Counting the number of characters

\small

```{r}
mystrings <- c("one fish", "two fish", "red fish", "blue fish")
nchar(mystrings)
```

## The ``stringr` package

\small

- Character string manipulation in R has evolved as
a collection of tools whose name and function is 
taken from string manipulation tools in Unix and
other scripting languages such as Perl.
- The `stringr` package aims for a cleaner interface
for tasks that relate to detecting, extracting, replacing
and splitting on substrings.
- Will indicate that a function is from stringr by 
prepending `stringr::` to the function name.
```{r}
library(stringr)
```


## Detecting substrings

\footnotesize

- `grep()` returns the indices of strings
that contain a substring,
while `grepl()` returns a logical vector:
```{r}
pattern <- "red"
grep(pattern,mystrings)
mystrings[grep(pattern,mystrings)]
grepl(pattern,mystrings)
mystrings[grepl(pattern,mystrings)]
```

## Detecting substrings with `stringr::str_detect()`

\small

- Works like `grepl()` but note that we *switch*
the order of the arguments:
```{r}
str_detect(mystrings,pattern)
mystrings[str_detect(mystrings,pattern)]
```

## Finding substring starting position

\footnotesize

- `regexpr()` returns the start of the first occurance
of a pattern, `gregexpr()` returns the start of 
all occurances.

\scriptsize

```{r}
Seuss <- paste(mystrings,collapse=", "); Seuss
regexpr("fish",Seuss) 
gregexpr("fish",Seuss)
```

## Finding substring starting position with `stringr`

\small

- `stringr` analogs to `regrexpr` and `gregexpr` are
`str_locate` and `str_locate_all`, with argument
order reversed.
```{r}
str_locate(Seuss,"fish")
str_locate_all(Seuss,"fish")
```


## Extracting substrings by start and stop position

\small

- We saw `substr()` in the example of lecture 3 where 
we read in purchase amounts and wanted to remove the
`$`. 
- Takes a character string, or vector of strings, as
argument. Specify start and stop character.
- Another example
```{r}
substr("this string has 30 characters!",start=10,stop=20) 
```

## `substr()` with big start and stop

\small

```{r}
bignum <- 1000
substr("this string has 30 characters!",start=10,stop=bignum) 
substr("this string has 30 characters!",start=31,stop=bignum) 
```

- If stop greater than number of characters, stop at the 
end of the string.
- If `start` greater than number of characters, return `""`

## Note: `substr` can do replacements to character variables

\small

But its use is not very intuitive:

```{r}
x<-"this string has 30 characters!"
substr(x,start=10,stop=20)  <- c("X")
x
substr(x,start=10,stop=20)  <- c("XXXXXXX")
x
substr(x,start=10,stop=20)  <- c("XXXXXXXXXXXXXXX")
x
```

## Replacing (substituting) substrings 

\small

- `sub()` and `gsub()` replace the first and all 
occurrences of a substring with a replacement.
```{r}
sub("fish","bird",Seuss)
gsub("fish","bird",Seuss)
```

## Replacing substrings with `stringr`

- Use `str_replace` and `str_replace_all`.
```{r}
str_replace(Seuss,"fish","bird") # replace first occurance
str_replace_all(Seuss,"fish","bird") # replace all
```



## Splitting strings with `strsplit`

\small

- `strsplit()` splits a vector of character 
strings on a specified separator and returns a list.
```{r}
mystrings <- c("this is a string", "so is this")
strsplit(mystrings,split=" ")
```

## `strsplit()` on special characters

\footnotesize

- Some characters, such as `.`, have special meaning when 
used as part of the `split` argument.
    - more on these special characters and "regular expressions" soon
- To match the `split` argument exactly, use
`fixed=TRUE`
```{r}
mystrings <- c("20.50", "33.33")
strsplit(mystrings,split=".")
strsplit(mystrings,split=".",fixed=TRUE)
```


## Splitting with `stringr`

\small

- The `str_split()` command is similar to `strsplit()`
```{r}
strsplit(mystrings,split=".")
strsplit(mystrings,split=".",fixed=TRUE)
```

## Combining strings with `paste()`

\small

- `paste()` glues together strings or vectors of strings separated by a user-specified separator (default " ").
    - The default separator of `paste0()` is no-space "".
```{r}
mystrings <- c("21.33","33.33")
paste(mystrings[1],mystrings[2])
paste("$",mystrings,sep="") 
```

- We can also paste together elements of a vector
```{r}
paste(mystrings,collapse=" ")
```



## Working with string patterns: regular expressions

\small 

- The string manipulations so far that involve
substrings have used fixed, or literal, substrings.
- Sometimes we would prefer to identify strings 
that match a pattern.
- A regular expression (abbreviated regex) is a string
of characters used to specify a search pattern
- Regular expressions is a complex topic. We'll only 
cover a simple case.
    
\scriptsize

- Learn more with the following references: 
    - RStudio Regular Expressions Cheatsheet:
\url{https://www.rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf}
    - Regular expressions secton of Prof. Bryan's Stat545 at UBC \url{http://stat545.com/block028_character-data.html}
    - The Strings chapter of R for Data Science \url{http://r4ds.had.co.nz/strings.html}


## A simple pattern with `.`

\small

- To illustrate pattern matching, use a simple pattern
`p.n`, meaning `p` followed
by any any character, followed by `n`.

```{r}
pattern <- "p.n"
mystrings <- c("pineapple","apple","pen")
```

## Detecting patterns

\small

- The functions `grep`, `grepl` and `str_detect` 
all accept regular expressions as the pattern to 
find; e.g.,
```{r}
str_detect(mystrings,pattern)
```

## Splitting on a pattern

\small

- `strsplit` and `str_split` accept regular expressions 
to split on; e.g.,
```{r}
str_split(mystrings,pattern)
```

## Locating a pattern

\small

- The string location functions `regexpr`, `gregexpr`,
`str_locate` and `str_locate_all` accept regular expressions;
e.g.,
```{r}
str_locate(mystrings,pattern)
```

## Extracting patterns

\small

- We previously extracted substrings based
on start and stop postition.
- Can also extract patterns.
```{r}
str_extract(mystrings,pattern)
str_match(mystrings,pattern)
```

## Replacing patterns

\small

- `sub`, `gsub`, `str_replace` and `str_replace_all`
accept regular expressions; e.g., 
```{r}
str_replace(mystrings,pattern,"PPAP")
```

- The replacement string is literal; e.g.,
```{r}
str_replace(mystrings,pattern,"p.n")
```

