---
title: "lecture 5"
author: "Brad McNeney"
date: '2018-02-01'
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Working with character strings

\small

* Fixed, or literal strings:
    - count the number of characters in a string
    - detect (yes/no) or find (starting position) substrings 
    - extract and substitute substrings
    - split and combine strings

* String patterns:
    - detect, find, extract and substitute

* Througout, illustrate "base" R utilities and those
from the `stringr` package

* A summary of what we discuss is available on the cheat sheet at
https://www.rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf
    




## The ``stringr` package

\small

- Character string manipulation in base R has evolved over time
as a bit of a patch-work of tools.
    * The names and functionality of these tools has
    been taken from string manipulation tools in Unix and
    scripting languages like Perl.
    * Not very familiar to non-Unix users.
- The `stringr` package aims for a cleaner interface
for tasks that relate to detecting, extracting, replacing
and splitting on substrings.

```{r}
library(stringr)
```

## Counting the number of characters

\small

```{r}
mystrings <- c("one fish", "two fish", "red fish", "blue fish")
nchar(mystrings)
```

## Detecting substrings

\footnotesize

- The base R function `grep()` returns the indices of strings
that contain a substring,
while `grepl()` returns a logical vector:
```{r}
pattern <- "red"
grep(pattern,mystrings)
mystrings[grep(pattern,mystrings)]
grepl(pattern,mystrings)
mystrings[grepl(pattern,mystrings)]
```

## Detecting substrings with `stringr::str_detect()`

\small

- Works like `grepl()` but note that we *switch*
the order of the arguments:
```{r}
str_detect(mystrings,pattern)
mystrings[str_detect(mystrings,pattern)]
```

## Finding substring starting position

\footnotesize

- The base R function `regexpr()` returns the start of the first occurance
of a pattern, `gregexpr()` returns the start of 
all occurances.
    * Also returned is an attribute `match.length`, which is the length
    of the matching string.
    * Also returned is an attribute `useBytes`, whose definition is 
    technical and which we will ignore.

\small

```{r}
Seuss <- paste(mystrings,collapse=", "); Seuss
regexpr("fish",Seuss) 
```

##

\small

```{r}
gregexpr("fish",Seuss)
```

## Finding substring starting position with `stringr`

\small

- `stringr` analogs to `regrexpr` and `gregexpr` are
`str_locate` and `str_locate_all`, with argument
order reversed.
```{r}
str_locate(Seuss,"fish")
str_locate_all(Seuss,"fish")
```


## Extracting substrings by start and stop position

\small

- We saw `substr()` in the example of lecture 3 where 
we read in purchase amounts and wanted to remove the
`$`. 
- Takes a character string, or vector of strings, as
argument. Specify start and stop character.
- Another example
```{r}
substr("this string has 30 characters!",start=10,stop=20) 
```

## `substr()` with big start and stop

\small

```{r}
bignum <- 1000
substr("this string has 30 characters!",start=10,stop=bignum) 
substr("this string has 30 characters!",start=31,stop=bignum) 
```

- If stop greater than number of characters, stop at the 
end of the string.
- If `start` greater than number of characters, return `""`

## Note: `substr` can do replacements to character variables

\small

But its use is not very intuitive:

```{r}
x<-"this string has 30 characters!"
substr(x,start=10,stop=20)  <- c("X") # Fewer than 11 in replacement
x
substr(x,start=10,stop=20)  <- c("XXXXXXX") # Fewer than 11
x
substr(x,start=10,stop=20)  <- c("XXXXXXXXXXXXXXX") # More than 11
x
```

## Replacing (substituting) substrings 

\small

- `sub()` and `gsub()` replace the first and all 
occurrences of a substring with a replacement, respectively.
```{r}
sub("fish","bird",Seuss)
gsub("fish","bird",Seuss)
```

## Replacing substrings with `stringr`

- Use `str_replace` and `str_replace_all`.
```{r}
str_replace(Seuss,"fish","bird") # replace first occurance
str_replace_all(Seuss,"fish","bird") # replace all
```



## Splitting strings with `strsplit`

\small

- `strsplit()` splits a vector of character 
strings on a specified separator and returns a list with one
list element per vector element.
```{r}
mystrings <- c("this is a string", "so is this")
strsplit(mystrings,split=" ")
```

## `strsplit()` on special characters

\footnotesize

- Some characters, such as `.`, have special meaning when 
used as part of the `split` argument.
    - more on these special characters and "regular expressions" soon
- To match the `split` argument exactly, use
`fixed=TRUE`
```{r}
mystrings <- c("20.50", "33.33")
strsplit(mystrings,split=".") # Splits on each of the 5 chars
strsplit(mystrings,split=".",fixed=TRUE)
```


## Splitting with `stringr`

\small

- The `str_split()` command is similar to `strsplit()`,
but with argument `pattern` instead of `split`.
    - wrap pattern in `fixed()` for a fixed string
```{r}
str_split(mystrings,pattern=".")
str_split(mystrings,pattern=fixed("."))
```

## Combining strings with `paste()`

\small

- `paste()` glues together strings or vectors of strings separated by a user-specified separator (default " ").
    - The default separator of `paste0()` is no-space "".
```{r}
mystrings <- c("21.33","33.33")
paste(mystrings[1],mystrings[2])
paste("$",mystrings,sep="") 
```

- We can also paste together elements of a vector
```{r}
paste(mystrings,collapse=" ")
```



## Working with string patterns: regular expressions

\small 

- The string manipulations so far that involve
substrings have used fixed, or literal, substrings.
- Sometimes we would prefer to identify strings 
that match a pattern.
- A regular expression (abbreviated regex) is a string
of characters used to specify a search pattern
- Regular expressions is a complex topic. We'll only 
cover the basics.
    
\scriptsize

- Learn more with the following references: 
    - RStudio Regular Expressions Cheatsheet:
\url{https://www.rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf}
    - Regular expressions secton of Prof. Bryan's Stat545 at UBC \url{http://stat545.com/block028_character-data.html}
    - The Strings chapter of R for Data Science \url{http://r4ds.had.co.nz/strings.html}


## A simple pattern with `.`

\small

- To illustrate pattern matching, use a simple pattern
`p.n`, meaning `p` followed
by any any character, followed by `n`.

```{r}
pattern <- "p.n"
mystrings <- c("pineapple","apple","pen")
```

## Detecting patterns

\small

- The functions `grep`, `grepl` and `str_detect` 
all accept regular expressions as the pattern to 
find; e.g.,
```{r}
str_detect(mystrings,pattern)
```

## Splitting on a pattern

\small

- `strsplit` and `str_split` accept regular expressions 
to split on; e.g.,
```{r}
str_split(mystrings,pattern)
```

## Locating a pattern

\small

- The string location functions `regexpr`, `gregexpr`,
`str_locate` and `str_locate_all` accept regular expressions;
e.g.,
```{r}
str_locate(mystrings,pattern)
```

## Extracting patterns

\small

- We previously extracted substrings based
on start and stop postition.
- Can also extract patterns.
```{r}
str_extract(mystrings,pattern)
str_match(mystrings,pattern)
```

## Replacing patterns

\small

- `sub`, `gsub`, `str_replace` and `str_replace_all`
accept regular expressions; e.g., 
```{r}
str_replace(mystrings,pattern,"PPAP")
```

- The replacement string is literal; e.g.,
```{r}
str_replace(mystrings,pattern,"p.n")
```


## Adding `*` and `+` quantifiers to `.`

\small

- The combinations `.*` and `.+` match 
multiple characters.
    - E.G., `f.*n` matches `f` followed by 0 or 
    more characters, followed by `n`.
    - `f.+n` matches `f` followed by 1 or more 
    characters, followed by `n`.
```{r}
mystrings <- c("fun","for fun","fn")
pattern1 <- "f.*n"; pattern2 <- "f.+n"
str_extract(mystrings,pattern1)
str_extract(mystrings,pattern2)
```

## "Greedy" matching with `*`

* The `*` quantifier matches the longest possible string.

```{r}
mystrings <- c("fun","fun, fun, fun","fn")
pattern1 <- "f.*n"
str_extract(mystrings,pattern1)
```

## Numerical quantifiers

* Use `{n}` to require exactly `n` matches

```{r}
pattern3 <- "f.{6}n"
str_extract(mystrings,pattern3)
```


## Other characters to match

* We have illustrated character matching on the pattern `.`, which is
any character.
* Instead we can specify a class of characters to match.


\small

```{r}
pattern4 <- "f[aeiou]*n" 
mystrings <- c("fan","fin","fun","fan, fin, fun",
               "friend","faint")
str_extract(mystrings,pattern4) 
```

##

\small

```{r}
str_extract_all(mystrings,pattern4)
```

