---
title: "Lecture 6"
author: "Brad McNeney"
date: '2017-02-07'
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Databases

## Relational database

- A relational database is a collection of 
tables. 
    - As statisticians, we think of rows 
    (a.k.a. records) as 
    sampled units, and columns as the variables
    measured on the units.
- Rows of one table are related to rows of 
others by keys
    - Each table has a key (primary key) that 
    uniquely identifies the rows.
    - Including the key of another table 
    (a foreign key)
    allows us to link (relate) records of 
    the two tables.
    
    
## Simple example database

\footnotesize

Example from [http://www.itl.nist.gov/div897/ctg/dm/sql_examples.htm]

- `STATION` table

ID  | City    | State  | Lat_N  | Long_W
--- | ------- | ------ | ------ | -------
13  | Phoenix | AZ     | 33     | 112
44  | Denver  | CO     | 40     | 105
66  | Caribou | ME     | 47     | 68

- `STATS` table (primary key suppressed; ID is 
foreign key)

ID  | Month | Temp_F | Rain_I
--- | ----- | ------ | ------
13  | 1     | 57.4   | 0.31
13  | 7     | 91.7   | 5.15
44  | 1     | 27.3   | 0.18
44  | 7     | 74.8   | 2.11
66  | 1     | 6.7    | 2.1
66  | 7     | 65.8   | 4.52

## Database software terminology

- Software refered to as relational database 
management systems (RDBMS).
    - Example implementations: MySQL, SQLite
- The language for querying the database is 
structured query language (SQL).
    - Similar implementations in all major RDBMS
- The database resides on a server, which
we access with a client.
    - Large databases will be on special-purpose
    remote servers.
    - Client runs on your computer.
    - However, client/server model usually
    involves layers of security
    that make access difficult.
    - We'll work mostly with SQLite -- server
    runs on your computer.
 
## Why databases in R?

- The data is stored in a RDBMS.
    - Our focus today
- Work with datasets that are too large to be 
stored in memory
    - Discuss later
    
## SQLite databases in R 

\small

- The `RSQLite` package contains an SQLite client
_and_ server we can use from R
    - SQLite is an open-source RDBMS engine bundled
    with RSQLite
    - Interface 
    with DB engine (connect, write to, etc.) _via_ 
    a common interface called `DBI`
    - Install `RSQLite` and `DBI` before starting
    - Load `DBI`; access functions from `RSQLite`
    with the `::` operator.
```{r}
library(DBI)
mydb <- dbConnect(RSQLite::SQLite(), "my-db.sqlite")
dbDisconnect(mydb)
#> [1] TRUE
unlink("my-db.sqlite")
```

## Notes

- The first line, 
`mydb <- dbConnect(RSQLite::SQLite(), "my-db.sqlite")`,
creates a "connection" to an SQLite database stored in the file "my-db.sqlite"
    - `SQLite()` is called the "driver". It 
    handles all the RDBMS-specific details of
    how the client and server communicate.
    - Initially the database is empty
- The second line, `dbDisconnect(mydb)`,
disconnects from the database, but does not
remove it.
- Remove with `unlink()` (a base R command).
- `dbDisconnect()` and all other commands
we will use of the 
form `db*` are generics from `DBI`

## Creating `STATION` and `STATS` tables

\scriptsize

```{r}
wdb <- dbConnect(RSQLite::SQLite(), "wdb.sqlite")
STATION <- data.frame(ID=c(13,44,66),
    City = c("Phoenix","Denver","Caribou"),
    State = c("AZ","CO","ME"),
    Lat_N = c(33,40,47),
    Long_W = c(112,105,68))
STATS <- data.frame(row = 1:6,
    ID = c(13,13,44,44,66,66),
    Month = c(1,7,1,7,1,7),
    Temp_F = c(57.4,91.7,27.3,74.8,6.7,65.8),
    Rain_I = c(0.31,5.15,0.18,2.11,2.1,4.52))
dbWriteTable(wdb,name='STATION', value = STATION, overwrite=TRUE) 
dbWriteTable(wdb,name='STATS', value = STATS, overwrite=TRUE)
dbListTables(wdb)
```

## Notes

- The `overwrite=TRUE` argument to the 
`dbWriteTable()` commands is necessary for
this demo, to allow re-knitting the document,
but is not necessary in general.
- Now that we have an example database,
we can see how to use SQL to extract data.
- Will interleave SQL tutorial and `DBI`. 

## SQL: Retrieve data from a single table



- The SQL `SELECT` statements:
    - `SELECT * FROM STATION;`
    - `SELECT City, Lat_N FROM STATION;`
    - `SELECT Month, Temp_F, Rain_I FROM STATS;`
- Filtering with the `WHERE` clause:
    - `SELECT * from STATION WHERE Lat_N >= 40;`
- Trailing `;` required for most SQL clients,
but not R `DBI`.

## DBI: Example data retrieval.

\scriptsize

```{r}
dbGetQuery(wdb,"SELECT * from STATION")
dbGetQuery(wdb,"SELECT City, Lat_N from STATION")
dbGetQuery(wdb,"SELECT * from STATION WHERE Lat_N>=40")
```

## DBI: Notes on `dbGetQuery()`

- `dbGetQuery()` calls three functions: 
    (i) `dbSendQuery()` sends the query to the DB, 
    (ii) `dbFetch()` fetches the "result set", and 
    (iii) `dbClearResult()` frees memory and 
    other resources associated with the result set.
- If the result set is too large to fit in memory,
you can split the fetching into batches.

## DBI: Batched queries 

\footnotesize

```{r}
rs <- dbSendQuery(wdb,"SELECT * FROM STATS")
while (!dbHasCompleted(rs)) {
  df <- dbFetch(rs, n = 2) # use n to set size of subset
  print(df)
}
dbClearResult(rs)
```

## DBI: Parametrized queries

\footnotesize

- Can pass the same query with several different
values of a parameter `x`.
    - Bind a value to the parameter with `dbBind()`
    
```{r}
rs <- dbSendQuery(wdb,"SELECT * FROM STATION WHERE Lat_N >= :x")
dbBind(rs,param = list(x=40))
dbFetch(rs)
dbBind(rs,param=list(x=45))
dbFetch(rs)
dbClearResult(rs)
```


## SQL: Joining tables

\small

- The purpose of related tables is to 
reduce redundancy.
    - For example, all the info on the stations
    appears once in the `STATION` table, and 
    need not be repeated in the `STATS` table.
- But what if we need the info on the stations
and the weather data? Need to "join" tables.
- Simplest join (inner join):
`SELECT * from STATION, STATS WHERE STATION.ID=STATS.ID`

## DBI: Example joins

\scriptsize

```{r}
dbGetQuery(wdb,"SELECT * from STATION, STATS WHERE STATION.ID=STATS.ID")
queryp1 <- "SELECT City, State, Month, Rain_I from STATION, STATS"
queryp2 <- "WHERE STATION.ID=STATS.ID AND Month = 1 AND Lat_N >= 40"
dbGetQuery(wdb,paste(queryp1,queryp2))
```


