---
title: 'Statistics 341:  Intro to Stat Computing and Exploratory Data Analysis with R'
subtitle: 'Lecture 1: Introduction to R and Getting Started'
author: "Brad Mcneney"
date: '2019-01-03'
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---
# Introducing R

## What is R?

- R is an open-source environment for statistical computing and graphics.
    - An implementation of the S language
    (<https://en.wikipedia.org/wiki/S_(programming_language)>)
- Started in the mid-1990's by Ross Ihaka and Robert Gentlemen at Auckland University
- Now maintained by a team of experts called the R Development Core Team
- A "packages" system allows any user to bundle R code, data and examples together.
- R and R packages are distributed through the Comprehensive R Archive Network (CRAN). 
- SFU has a CRAN mirror at <http://cran.stat.sfu.ca>

## What does "environment" mean?
- R is a fully-functioning programming environment with all the usual constructs, such as 
    - conditionals (if-then-else), 
    - loops 
    - user-defined functions.
- In addition there are built-in facilities for 
    - data input, storage, manipulation, and output
    - optimization, matrix computation, etc., 
    - random number generation,
    - data analysis and graphics.
- "Base" R is good, but it is the package system that makes R great.

## R packages
- The R package system is the key to R's success
- It has allowed statisticians and other data scientists to implement and distribute their work to be used by others.
- The R package system enforces some rules about how packages are structured, but differences in programming styles of package authors mean different interfaces
    - Users need to be aware of different data structures for input, output and of different styles of graphics
    
# Course objectives

## Course objectives

- Understand basic R data structures and programming
- Learn how to use base R and R package functions for data management, exploration, presentation and analysis
- Learn how to use packages from the "tidyverse", a collection of modern tools for data science.
    * <https://www.tidyverse.org/>

# Getting started

## Getting started with R, RStudio and git

- A brief set of instructions are on the canvas page 
<https://canvas.sfu.ca/courses/37593/pages/getting-started-with-r-rstudio-and-git>
- Please try to get R and RStudio installed and 
create an RStudio project linked to the class GitHub 
repository as soon as possible.
- Those still having trouble after the weekend should
attend one of the Wednesday tutorials (12:30-3:20)
for help, even if these are not your tutorial times.

## R reference cards

- RStudio has created several useful reference 
cards, called cheat sheets, and have collected
cheat sheets from other R users.
- See
<https://www.rstudio.com/resources/cheatsheets/>
- For getting started with RStudio you might find
the following helpful:
<https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf>

## Starting R

- Start R by starting RStudio. 
- The "Console" window is where you can type your commands.
- However, it is good practice to open an R script, type your
commands in the script, and then submit the commands to the R console.
    - `Session -> Set Working Directory` to set the working directory
    - `File -> New File -> R Script` to open a new R script
    - type your commands into the script
    - put your cursor on the line you want to submit and hit `Ctrl-enter`
- Save your script for later use.
- More on the RStudio interface at <https://support.rstudio.com/hc/en-us/sections/200107586-Using-RStudio>

# Quick-start R 

## R as a calculator
\small
```{r}
2+4
5*6
x<-6
x*5
```
- The assignment operator `<-` assigns the value 6 to the variable x


## Random number generation
\footnotesize
```{r}
set.seed(123) # optional, but makes computation reproducible
n<-100
y<-rnorm(n,sd=1)
hist(y)
```

## R data 

\footnotesize
```{r}
data(cars) # load a built-in example dataset, help(cars) for info
head(cars) # look at the first few rows
summary(cars) # summarize the variables
```

## Plotting data
\small
```{r}
plot(cars)
```

## Modelling
\footnotesize
```{r}
fit <- lm(dist~speed,data=cars) # fit a linear model, help(lm)
summary(fit)
```

## Modelling, continued
\small
```{r}
plot(cars)
abline(fit)
```

## Extracting model components 
\footnotesize
```{r}
head(residuals(fit))
vcov(fit)
```
- See the help page for more components

## Model diagnostics
```{r}
par(mfrow=c(2,2))
plot(fit)
```

## Reading 
- R and RStudio help pages/reference cards mentioned in this lecture

