---
title: "Lecture 6"
author: "Brad McNeney"
date: '2017-02-07'
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Databases

## Relational database

- A relational database is a collection of 
tables. 
    - As statisticians, we think of rows 
    (a.k.a. records) as 
    sampled units, and columns as the variables
    measured on the units.
- Rows of one table are related to rows of 
others by keys
    - Each table has a key (primary key) that 
    uniquely identifies the rows.
    - Including the key of another table 
    (a foreign key)
    allows us to link (relate) records of 
    the two tables.
    
    
## Simple example database

\footnotesize

Example from [http://www.itl.nist.gov/div897/ctg/dm/sql_examples.htm]

- `STATION` table

ID  | City    | State  | Lat_N  | Long_W
--- | ------- | ------ | ------ | -------
13  | Phoenix | AZ     | 33     | 112
44  | Denver  | CO     | 40     | 105
66  | Caribou | ME     | 47     | 68

- `STATS` table (primary key suppressed; ID is 
foreign key)

ID  | Month | Temp_F | Rain_I
--- | ----- | ------ | ------
13  | 1     | 57.4   | 0.31
13  | 7     | 91.7   | 5.15
44  | 1     | 27.3   | 0.18
44  | 7     | 74.8   | 2.11
66  | 1     | 6.7    | 2.1
66  | 7     | 65.8   | 4.52

## Database software terminology

- Software refered to as relational database 
management systems (RDBMS).
    - Example implementations: MySQL, SQLite
- The language for querying the database is 
structured query language (SQL).
    - Similar implementations in all major RDBMS
- The database resides on a server, which
we access with a client.
    - Large databases will be on special-purpose
    remote servers.
    - Client runs on your computer.
    - However, client/server model usually
    involves layers of security
    that make access difficult.
    - We'll work mostly with SQLite -- server
    runs on your computer.
 
## Why databases in R?

- The data is stored in a RDBMS.
    - Our focus today
- Work with datasets that are too large to be 
stored in memory
    - Discuss later
    
## SQLite databases in R 

\small

- The `RSQLite` package contains an SQLite client
_and_ server we can use from R
    - SQLite is an open-source RDBMS engine bundled
    with RSQLite
    - Interface 
    with DB engine (connect, write to, etc.) _via_ 
    a common interface called `DBI`
    - Install `RSQLite` and `DBI` before starting
    - Load `DBI`; access functions from `RSQLite`
    with the `::` operator.
```{r}
library(DBI)
mydb <- dbConnect(RSQLite::SQLite(), "my-db.sqlite")
dbDisconnect(mydb)
#> [1] TRUE
unlink("my-db.sqlite")
```

## Notes

- The first line, 
`mydb <- dbConnect(RSQLite::SQLite(), "my-db.sqlite")`,
creates a "connection" to an SQLite database stored in the file "my-db.sqlite"
    - `SQLite()` is called the "driver". It 
    handles all the RDBMS-specific details of
    how the client and server communicate.
    - The following argument is the file name for the 
    database. Other drivers need details like
    username and password to connect.
    - Initially the database is empty
- The second line, `dbDisconnect(mydb)`,
disconnects from the database, but does not
remove it.
- Remove with `unlink()` (a base R command).
- `dbDisconnect()` and all other commands
we will use of the 
form `db*` are generics from `DBI`

## Aside: MySQL database driver in R

- Wasn't able to get easy access to a MySQL database
on campus. 
- In case it is of later use, here are the arguments
to `dbConnec()` for connecting to a password-protected
database for Stat 341.

\footnotesize

```
con <- dbConnect(MySQL(), user='stataccess', 
                  password='<password goes here>', 
                  dbname='stat341', 
                  host='muncho.its.sfu.ca')
dbListTables(con)
```
## Creating `STATION` and `STATS` tables

\scriptsize

```{r}
wdb <- dbConnect(RSQLite::SQLite(), "wdb.sqlite")
STATION <- data.frame(ID=c(13,44,66),
    City = c("Phoenix","Denver","Caribou"),
    State = c("AZ","CO","ME"),
    Lat_N = c(33,40,47),
    Long_W = c(112,105,68))
STATS <- data.frame(row = 1:6,
    ID = c(13,13,44,44,66,66),
    Month = c(1,7,1,7,1,7),
    Temp_F = c(57.4,91.7,27.3,74.8,6.7,65.8),
    Rain_I = c(0.31,5.15,0.18,2.11,2.1,4.52))
dbWriteTable(wdb,name='STATION', value = STATION, overwrite=TRUE) 
dbWriteTable(wdb,name='STATS', value = STATS, overwrite=TRUE)
dbListTables(wdb)
```

## Notes

- The `overwrite=TRUE` argument to the 
`dbWriteTable()` commands is necessary for
this demo, to allow re-knitting the document,
but is not necessary in general.
- Now that we have an example database,
we can see how to use SQL to extract data.
- Will interleave SQL tutorial and `DBI`. 

## SQL: Retrieve data from a single table



- The SQL `SELECT` statements:
    - `SELECT * FROM STATION;`
    - `SELECT City, Lat_N FROM STATION;`
    - `SELECT Month, Temp_F, Rain_I FROM STATS;`
- Filtering with the `WHERE` clause:
    - `SELECT * from STATION WHERE Lat_N >= 40;`
- Trailing `;` required for most SQL clients,
but not R `DBI`.

## DBI: Example data retrieval.

\scriptsize

```{r}
dbGetQuery(wdb,"SELECT * from STATION")
dbGetQuery(wdb,"SELECT City, Lat_N from STATION")
dbGetQuery(wdb,"SELECT * from STATION WHERE Lat_N>=40")
```

## DBI: Notes on `dbGetQuery()`

- `dbGetQuery()` calls three functions: 
    (i) `dbSendQuery()` sends the query to the DB, 
    (ii) `dbFetch()` fetches the "result set", and 
    (iii) `dbClearResult()` frees memory and 
    other resources associated with the result set.
- If the result set is too large to fit in memory,
you can split the fetching into batches.

## DBI: Batched queries 

\footnotesize

```{r}
rs <- dbSendQuery(wdb,"SELECT * FROM STATS")
while (!dbHasCompleted(rs)) {
  df <- dbFetch(rs, n = 2) # use n to set size of subset
  print(df)
}
dbClearResult(rs)
```

## DBI: Parametrized queries

\footnotesize

- Can pass the same query with several different
values of a parameter `x`.
    - Bind a value to the parameter with `dbBind()`
    
```{r}
rs <- dbSendQuery(wdb,"SELECT * FROM STATION WHERE Lat_N >= :x")
dbBind(rs,param = list(x=40))
dbFetch(rs)
dbBind(rs,param=list(x=45))
dbFetch(rs)
dbClearResult(rs)
```


## SQL: Joining tables

\small

- The purpose of related tables is to 
reduce redundancy.
    - For example, all the info on the stations
    appears once in the `STATION` table, and 
    need not be repeated in the `STATS` table.
- But what if we need the info on the stations
and the weather data? Need to "join" tables.
- Simplest join (inner join):
`SELECT * from STATION, STATS WHERE STATION.ID=STATS.ID`

## DBI: Example joins

\scriptsize

```{r}
dbGetQuery(wdb,"SELECT * from STATION, STATS WHERE STATION.ID=STATS.ID")
queryp1 <- "SELECT City, State, Month, Rain_I from STATION, STATS"
queryp2 <- "WHERE STATION.ID=STATS.ID AND Month = 1 AND Lat_N >= 40"
dbGetQuery(wdb,paste(queryp1,queryp2))
```

## SQL: Left joins

\small

- The inner join returns data for cities
in **both** the `STATION` and `STATS` tables.
- If we want all stations, use
a left join.
    - First add another station with no data in `STATS`
    - Miami, FL is at Lat 26 and Long 80.
    - Give Miami station ID 77.
    - In SQL we'd add Miami and do the left join with 
```
INSERT INTO STATION VALUE (77,'Miami','FL',26,80)
SELECT * FROM STATION 
LEFT JOIN STATS ON STATION.ID = STATS.ID
```

- Many other types of SQL joins. See
[https://www.tutorialspoint.com/sqlite/sqlite_using_joins.htm]
for a summary of joins in SQLite.

## DBI: Adding to a table and left join

\footnotesize

```{r}
miami <- data.frame(ID=77,City="Miami",State="FL",Lat_N=26,Long_W=80)
dbWriteTable(wdb,name='STATION', value = miami, append=TRUE)
qq<-"SELECT * FROM STATION LEFT JOIN STATS ON STATION.ID = STATS.ID"
dbGetQuery(wdb,qq)
```

- Note: `append=TRUE` adds to the current table

## SQL: Table indices

\small

- A query like
`SELECT * FROM STATION WHERE Lat_N >= 40`
requires that the RDBMS read the `Lat_N` 
value in every row of `STATION` and 
return the rows where `Lat_N` is 40 or more
- Such a query can be made much faster 
by creating an "index" on `Lat_N`.
    - An index is a table in the database, sorted
    on the indexed variable.
    - See [http://www.sqlite.org/queryplanner.html] for
    a nice description of how indexing columns speeds
    up searches.
    
## DBI: Create an index with `dbExecute()`

\small

- Use `dbExecute()` to execute queries that
do not return tabular data.

```{r}
dbExecute(wdb,"CREATE INDEX indx ON STATION(Lat_N)")
```


## Exercises


1. Select the stations from the state of Colorado
2. Add Vancouver as 
a new station in the `STATION` table
(Use BC as the "State" and the latitude 49 and
longitude 123)
3. Do an inner join that returns a table with city,
state, and temperatures for July from 
cities at north latitude 40 or less.
4. Repeat (3) as a left join.


## Clean up

```{r}
dbDisconnect(wdb)
unlink("wdb.sqlite")
```


# Merging data frames

## Inner joining data frames with `merge()` 

\footnotesize

- The `merge()` function in R does 
SQL-like joins on data frames.
    
```{r}
STATION <- rbind(STATION,miami)
merge(STATION,STATS,by="ID")
```

## Left joining data frames with `merge()` 

\footnotesize

```{r}
merge(STATION,STATS,by="ID",all.x=TRUE)
```

## Join functions in `dplyr`

\footnotesize

- In `dplyr` the functions for joining are
more explicitly named

```{r, warning=FALSE,message=FALSE}
library(dplyr)
inner_join(STATION,STATS,by="ID")
```

## Left join function in `dplyr`

\footnotesize

```{r}
left_join(STATION,STATS,by="ID")
```

## Exercises with `dplyr`

1. Select the stations from the state of Colorado
(`filter()`)
2. Select the City and State from the stations
using the `select()` command from `dplyr`:
`select(STATION,City,State)`1
3. Add Vancouver as 
a new station in the `STATION` data frame.
Use BC as the "State" and the latitude 49 and
longitude 123. (`rbind()` or `bind_rows()`)
4. Do an inner join that returns a table with city,
state, and temperatures for July from 
cities at north latitude 40 or less.
(`inner_join()` and `select()`)
5. Repeat (3) as a left join.
(`left_join()` and `select()`)


