---
title: "lecture 4"
author: "Brad McNeney"
date: '2018-01-25'
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```

# Data management

## Data management overview

- After reading data into a data frame, we need to manage it:
    - print, view and edit data frames, add/delete variables
    - derive new variables from old
    - merge and reshape datasets
- There are many tools in "base" R.
- Many more in `dplyr`.
    - Will discuss `dplyr` alternatives where possible.
    
# Print, view, edit
    
## `print()`, `View()` and `edit()`

- `print()` prints R objects
    - This function is "generic", meaning that 
    it will try to find the 
    specific function to print specific objects
    (e.g., `print.data.frame`).
- `View()` launches a new window (or RStudio tab)
to view a data frame and `edit()` launches a 
data editor.

##

\small

```{r}
testdf = data.frame(ID=1:3,age=c(8,11,14),height=c(52,63,70))
print(testdf) # calls print.data.frame(testdf)
# View(testdf)
# edit(testdf)
```


## Access variables in a data frame

\small

- Can use what we learned about subsetting:
```{r}
testdf$ratio <- testdf$height/testdf$age
```

- or can use `with()`
```{r}
testdf$ratio <- with(testdf,height/age)
```

- Notice how we can add a new variable to `testdf` by 
assignment. 

## Using `attach()` to attach a data frame

\small

- What `with()` is doing is (i) create a 
temporary environment, (ii) copy the variables in `testdf` into this environment, (iii) evaluate the expression `height/age` in this temporary environment and (iv) return the results.
- We can do this manually with `attach(testdf)`
followed by the expression/assignment, and 
then use `detach("testdf")` to get rid of the 
temporary environment.
- However, the original data frame and its copy can get out
of sync and cause confusion, or we might forget
to `detach()`.
- Attaching data frames is generally frowned upon.


## Adding and deleting variables from a data frame

\small

- We saw how `$` can be used to add variables to 
a data frame. 
- Remove variables by setting to `NULL`
```{r}
testdf$ratio <- NULL
testdf
```

- To rename a variable, can add it under new name
and remove old variable. 

## Renaming with `rename()`

* The `rename()` function in the `dplyr` package
can be used to rename variables.

\small

```{r}
library(dplyr)
testdf <- rename(testdf,A_G_E=age)
testdf
testdf <- rename(testdf,age = A_G_E) # set back
```


# Derived variables 

## Adding derived variables with `tranform()` 

\small

- `transform()` together with an assignment can add
and/or modify variables:
```{r}
testdf <- transform(testdf,ID = c("E","K","H"),
                    ratio = height/age)
testdf
```

## Adding derived variables with `within()` 

- `within()` is similar to `transform()` but allows us to use 
variables we create in the call:
```{r}
testdf <- within(testdf, {
  heightcm <- height*2.54 # now use new variable heightcm
  ratiocm <- heightcm/age
  })
testdf
```

## Adding derived variables with `mutate()`

\small

- `mutate()` from the `dplyr` package is very similar to
`transform()`
    - Being from `dplyr` suggests using the forward 
    pipe %>% to chain multiple `mutate()`s
```{r, warning=FALSE,message=FALSE}
testdf %>%
  select(ID,age,height) %>% 
  mutate(heightcm = height*2.54) %>%
  mutate(ratiocm = heightcm/age) -> testdf
testdf
```

- Notice the use of `->` to assign the results of our
data manipulations.

## Creating and working with categorical variables

- We may want to 
    - create categorical by binning a numeric
    - create categorical with logical conditions
    - recode categories
    
## Binning a numeric variable with `cut()`

\small 

- Creates a factor based on equal-width bins by default:
```{r}
set.seed(1)
n <- 100
age <- sample(17:85,size=n,replace=TRUE)
agecat <- cut(age,breaks=5)
table(agecat)
```

## DIY bins with `cut()`

\small

- Custom bins. Be careful not not to exclude 
any data values.
```{r}
agecat <- cut(age,breaks = c(17,30,40,50,60,70,80))
# print agecat to see <NA>s
agecat[age==17]
agecat <- cut(age,breaks = c(15,30,40,50,60,70,80))
table(agecat)
```

## Create categorical from logical conditions

\small

- Usual strategy is to initialize a vector to a baseline
category and then use logical conditions to 
assign category of subsets.
```{r}
group <- sample(c("A","B"),size=n,replace=TRUE)
# print (cbind(age,group))
catvar <- rep(1,n) 
catvar[age<50 & group=="A"] <- 2
# print (cbind(age,group))
catvar[age<60 & group=="B"] <- 3
table(catvar)
```

## Recoding variables

\footnotesize

- For numeric or character categories use logical conditions.
```{r}
catvar[catvar==3] <- 11 # 11 gets recycled
```

- For factors, remember that they are numeric with character
labels, or levels -- just change the levels
```{r}
head(agecat)
levels(agecat)[1] <- "[17,30]"
head(agecat)
```

## Using `recode()` and `recode_factor()` from `dplyr`

\footnotesize
- Can recode multiple values at once and use with `%>%`
```{r}
# Enclose numeric values in backticks
head(recode(catvar,`1`="pen",`2`="pineapple",`11`="apple"))

cut(age,breaks = c(15,30,40,50,60,70,80,90)) %>%
  recode_factor("(15,30]" = "[17,30]", "(80,90]" = "(80,100]") %>% 
  head()
```

- Notice how the order of the levels has changed.

## Dates
 
\small

- We have seen the `as.Date()` function for coercing 
character strings to `Date` objects.
    - The function first tries the format yyyy-mm-dd,
    then yyyy/mm/dd. 
- Summary functions such as `mean()` and `diff()` can handle
`Date` objects.
```{r}
dd <- c("2002-04-02","2005-08-17","2008-08-12")
dd <- as.Date(dd)
mean(dd)
diff(dd)
```

## Reading dates in other formats

\small

- If your dates are in a format other than yyyy-mm-dd
or yyyy/mm/dd you will 
have to specify. 
- The formatting rules are described in `help(strptime)`.
```{r}
dd <- c("05/14/1966","04/02/2002","08/17/2005","08/12/2008")
dd <- as.Date(dd,format = "%m/%d/%Y")
dd
mean(dd)
diff(dd)
```
