---
title: 'Lecture 8'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```

# Iterating

## Iterating
* Reference: Iteration chapter of R for Data Science by 
Wickham and Grolemund.
    - Published book: Chapter 17
    - Online book: Chapter 21: http://r4ds.had.co.nz/iteration.html

* The use of `lapply()` in the previous lecture is 
an example of iterating:
    - Our data is in a vector (list), and we want to perform
    the same operation on each element.
* Tools that we have discussed that are useful for iterating are 
`for()` and `while()` loops.
* Iteration is so common that special tools have been developed
with the aim of reducing the amount of code (and therefore
errors) required for common iterative tasks.
    - Tools in base R include the `apply()`
    family of functions. 
    - A tidyverse package called `purrr` includes more.

## Vectorizing to avoid iteration

* Before iterating, think about whether it is 
possible to use vectorization instead.
* Example:

\scriptsize
```{r}
f <- function(x) {
  n <- length(x); out <- vector(mode="numeric",length=n)
  for(i in 1:n) {
    if(x[i]>0) {
      out[i] <- log(x[i])
    } else {
      out[i] <- NA
    }
  }
  out
}
fvec <- function(x) {
  n <- length(x); out <- vector(mode="numeric",length=n)
  ind <- (x>0)
  out[ind] <- log(x[ind])
  out[!ind] <- NA # recycles NA
  out
}
```
    
## Example data

* To illustrate iteration we simulate data and fit four 
regression models.

\small

```{r}
set.seed(42)
n <- 100
x1 <- rnorm(n); x2<-rnorm(n)
y1 <- x1 + rnorm(n,sd=.5); y2 <- x1+x2+rnorm(n,sd=.5)
y3 <- x2 + rnorm(n,sd=.5); y4 <- rnorm(n,sd=.5)
rr <- list(fit1 = lm(y1 ~ x1+x2),
  fit2 = lm(y2 ~ x1+x2),
  fit3 = lm(y3 ~ x1+x2),
  fit4 = lm(y4 ~ x1+x2))
```

## Extracting the regression coefficient for `x1`

* Using a `for()` loop, we initialize an object to hold the 
**output**, 
loop along a **sequence** of values for an index variable, 
and execute the **body** for each value of the index variable.

\small

```{r}
beta1hat <- vector("double",length(rr))
for(i in seq_along(rr)) { # safer than 1:length(rr) 
  beta1hat[i] <- coefficients(rr[[i]])["x1"]
}
beta1hat
```

## Looping over elements of a set

* The index set in the `for()` loop can be general. 
    - We might use this generality to loop over named components of a list.
    
\small

```{r}
fits <- paste0("fit",1:4)
for(ff in fits) {
  print(coefficients(rr[[ff]])["x1"])
}
```

\normalsize

* Looping over a set makes it harder to save the results, though.

## The body of a loop can be a small part of the code

* In our examples, most of the code is for setting up the 
output and looping, with very little to do with the body.

* To illustrate, consider a small change: instead of the estimated coefficient of `x1` we wanted
the estimated coefficient of `x2`:

\small

```{r}
beta1hat <- vector("double",length(rr))
for(i in seq_along(rr)) { # safter than 1:length(rr) 
  beta1hat[i] <- coefficients(rr[[i]])["x2"]
}
beta1hat
```

## Using `lapply()`

* The intent of `lapply()` is to take care of 
the output and the loop, allowing us to focus 
on the body.

\scriptsize

```{r}
b1fun <- function(fit) { coefficients(fit)["x1"] } # body
bfun <- function(fit,cc) { coefficients(fit)[cc] } # body
lapply(rr,b1fun)  # or sapply(rr,b1fun) or unlist(lapply(rr,b1fun))
lapply(rr,bfun,"x2")  
```

## Iterating with the `map()` functions from `purrr`

* The `purrr` package provides a family of functions 
`map()`, `map_dbl()`, etc. that do the same thing
as `lapply()` but work better with other tidyverse functions.
    * `map()` returns a list, like `lapply()`.
    * `map_dbl()` returns a double vector, etc.

\small

```{r}
library(purrr)
map_dbl(rr,b1fun) # or rr %>% map_dbl(b1fun)
```


## Pipes and `map()` functions

* Suppose we want to record a model summary returned by 
the `summary()` function.
    * `summary()` is a generic function. When applied to
    an `lm()` object it computes regression summaries like
    standard errors and model R$^2$.

\small

```{r}
rr %>%
  map(summary) %>%
  map_dbl(function(ss) { ss$r.squared })
```

\normalsize

## 

* Notice that we can define a function on-the-fly in the
call to a `map()` function.

* `map()` functions have a short-cut for function
definitions.

\small

```{r}
rr %>%
  map(summary) %>%
  map_dbl(~.$r.squared)
```

\normalsize

* In `~.` read `~` as "define a function" and `.` as "argument to 
the function"
    - Comment: This is a little too terse for my tastes,
    but I mention it in case you see it in practice.

# Exploratory Data Analysis

## Topics

- Exploratory data analysis, with emphasis
on `ggplot2` graphics, using the gapminder data.
    - Suppose we want to use information on `continent`, `year`,
    `pop` and `gdpPercap` to predict `lifeExp`.
- Base R graphics _vs_ grid graphics
- Introduction to `ggplot2`

# Exploratory analyses

## Exploratory analyses

\small

- Univariate summaries, such as means/medians, sds/IQRs, histrograms, to examine distributions 
and identify possible measurement errors.
- Pair-wise correlations, to look for relationships between variables
- Pair-wise regression relationships and added-variable-plots
    - Trends over time deserve special attention
- Illustrate with the `gapminder` data set.

\small

```{r}
library(gapminder)
data(gapminder)
```
    
## Univariate Summaries

\small

- Different summaries are appropriate for categorical and quantitative
variables
    - Tabulate categorical variables
    - Five number summary for quantitative variables

\scriptsize

```{r}
summary(gapminder)
```

## Comments on summaries

- Observations in `pop` and `gdpPercap` differ
by orders of magnitude
    - May be more informative to consider transformations
    of these variables.
    - For example, a log-10 transformation: one-unit 
    differences correspond to 10-fold increases.
- Aside: Which country has per-capita GDP of \$113,523? Or more
generally, which observations are in, say, the top 0.1\%?

\scriptsize

```{r}
library(dplyr)
filter(gapminder,gdpPercap > quantile(gdpPercap,0.999))
```


## Unviariate summaries by grouping variable

\small

- May be of interest to do summaries of 
some variables stratified by a grouping 
variable.

\scriptsize

```{r}
oldops <- options(tibble.width=Inf, tibble.print_max=Inf)
gm_byContinent <- group_by(gapminder,continent)
summarize(gm_byContinent,min(lifeExp),median(lifeExp),
          IQR(lifeExp),mean(lifeExp),sd(lifeExp),max(lifeExp))
options(oldops)
```

## Boxplots

\small

- Graphical representation of the five number summary for 
grouped data

\scriptsize

```{r, fig.width=4, fig.height=2.5}
library(ggplot2)
ggplot(gapminder,aes(x=continent, y=lifeExp)) + geom_boxplot()
```

## Boxplots, cont.

\scriptsize

```{r, fig.width=4, fig.height=2.5}
ggplot(gapminder,aes(x=continent, y=gdpPercap)) + geom_boxplot() 
```

- Distribution of log-transformed data may be more informative.

## Boxplots, cont.

\scriptsize

```{r, fig.width=4, fig.height=2.5}
ggplot(gapminder,aes(x=continent, y=gdpPercap)) +
  coord_trans(y="log10") + geom_boxplot() 
```

## Adding transformed variables to a dataset

\small

- Above suggests we add log of `gdpPercap` to the dataset.
- A similar exploration of the `pop` variable suggests we
include log of `pop` too.
- Will use log-base-10.

\scriptsize

```{r}
gapminder <- mutate(gapminder, 
                    log10Pop = log10(pop), 
                    log10GdpPercap = log10(gdpPercap))
```


## Histograms

\small

- Shows the shape of distributions and 
can suggest possible outliers 
- Stacked histograms:

\scriptsize

```{r, fig.width=4, fig.height=2.5, message=FALSE}
ggplot(gapminder,aes(x=lifeExp, color=continent)) + geom_histogram()
```

## Histograms, continued

- Histograms in different plot panels, or "facets":

\scriptsize

```{r, fig.width=4, fig.height=2.5, message=FALSE}
ggplot(gapminder,aes(x=lifeExp, color=continent)) + 
  geom_histogram() + facet_grid(continent ~ .)
```

## Histograms of the explanatory variables

- May also be of interest 

\scriptsize

```{r, fig.width=4, fig.height=2.5, message=FALSE}
ggplot(gapminder,aes(x=log10GdpPercap, color=continent)) +
  geom_histogram() + facet_grid(continent ~ .)
```



## Pairwise Regression relationships

\small

- Though pairwise relationships don't tell the whole story,
they are a useful starting point.
- The `GGally` package provides the function `ggpairs()`
to facilitate this.
    - Can do all possible pairs
    of variables, but I find this too hard to read for more
    than three  variables.
    
## Pairwise plots
    
\scriptsize

```{r, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
library(GGally)
gm_sub <- select(gapminder,continent,lifeExp,log10GdpPercap)
ggpairs(gm_sub) # Cut and paste into console to see better
```

## Time trends

\small

- Can represent time series by lines.
- There are many time series in these data -- need to 
make each line slightly transparent to account for overplotting 

\scriptsize

```{r, fig.height=3, fig.width=4, message=FALSE}
ggplot(gapminder,aes(x=year,y=lifeExp,group=country)) +
  geom_line(alpha=0.2)
```

## Time trends, cont.

\scriptsize

- Can add a statistical summary, like medians at each time,
or a smoother.
- Can also add colours for different continents.

```{r, fig.height=3, fig.width=4, message=FALSE}
ggplot(gapminder,aes(x=year,y=lifeExp,group=country,
                     color=continent)) + geom_line(alpha=0.2) + 
  geom_smooth(aes(x=year,y=lifeExp,group=continent),se=FALSE)
```

# Base R graphics

## Base R graphics

\small

- Very serviceable graphics system capable of 
producing publication-quality graphs.
- Create graphics by calling functions that either 
produce complete plots or add to plots
- Like adding paint to a canvas

## Base R examples

\scriptsize

```{r, fig.width=4,fig.height=3}
with(gapminder,plot(log10GdpPercap,lifeExp)) # or plot(lifeExp ~ log10GdpPercap)
title(main="life expectance vs log10 GDP percapita")
abline(lm(lifeExp ~ log10GdpPercap,data=gapminder),col="red",lwd=3)
```

## Base R examples

\scriptsize

```{r, fig.height=3, fig.width=4}
with(gapminder,boxplot(split(lifeExp,continent)))
```

## Base R examples

\scriptsize

```{r, fig.height=3, fig.width=4}
with(gapminder,plot(density(lifeExp)))
```

## Base R graphics: Where to learn more

- Paul Murrell's book:
[https://www.stat.auckland.ac.nz/~paul/RG2e/]
- Ross Ihaka's lectures:
[https://www.stat.auckland.ac.nz/~ihaka/787/lectures-r-graphics.pdf]

# `grid` graphics

## `grid` graphics

\small

- `grid` graphics is a low-level graphics system that 
allows fine control of graphics elements
- Users can create multiple
graphics regions, or "viewports",
that are arranged on the graphics device or nested within 
one another.
- Graphical objects, or "grobs" can be created and 
drawn on these viewports (e.g., lines, shapes).
- Grobs can be editted (e.g., change color of lines) and 
re-drawn

## `grid` graphics: Where to learn more

- Paul Murrell's book:
[https://www.stat.auckland.ac.nz/~paul/RG2e/]


# `ggplot2` basics

## `ggplot2` 

- `ggplot2` is implemented in `grid` graphics
- The `g`'s stand for Grammar of Graphics. 
    - Like English grammar is the way in which words
    are put together to form sentences, a grammar of 
    graphics is a way to put together basic graphical 
    elements to make a graph.
- To understand the grammar we need to define the 
basic elements.
    - Start with definitions (in bold), some of which are 
    too abstract to be useful until we get into 
    details.
- `ggplot`s can be built in layers, comprised of 
**data** a **mapping**, a **geom** and optionally **stats**
- The layers are arranged and labelled on the graph
by **scale**s and **coord**s.
- The data can also be broken into subsets and displayed
in separate graphs by a **facet** specification.

## Components of a `ggplot`: data and mappings

\small

- We start with the **data** we want to visualize and
a **mapping**, or aesthetic, that describes how these
data map to attributes on the plot.

\scriptsize

```{r, fig.width=3, fig.height=2}
p <- ggplot(gapminder,aes(x=log10GdpPercap,y=lifeExp,color=continent))
```

- From the dataset `gapminder`, the variable 
`log10GdpPercap` will be mapped to y-coordinates,
`lifeExp` will be mapped to the x-coordinates,
and `continent` will be perceived as colours.

## Components of a `ggplot`: geometric objects (geoms)

\small

- Geometric objects or **geoms** are things like
points and lines that we see on the plot.

\scriptsize

```{r, fig.width=4, fig.height=3}
p2 <- p + geom_point(alpha=0.5) 
```

- alpha is the transparency aesthetic, between 0 and 1, best
applied directly to the geom.

## Components of a `ggplot`: statistical transformations

\small

- Statistical transformations or **stats** summarize  the data;
e.g., a scatterplot smoother

\scriptsize

```{r, fig.width=4, fig.height=2}
p2 + stat_smooth()
```


## Components of a `ggplot`: scales

\small

- The **scales** are mappings from the data to the graphics device
    - domain of `continent` is the five continents, range is
    the hexidecimal of the 
    five colors represented on the graph
    - domain of `lifeExp` is 23.599 to 82.603, range is [0,1],
    which `grid` converts to a 
    range of vertical pixels on the graph.
    - legends and axes provide the inverse mapping

\scriptsize

```{r, fig.width=4, fig.height=1.7}
p2
```


## Components of a `ggplot`: coodinate system

\small

- The coordinate system is another layer in how the data
get mapped to the graphics device. 
    - Usually Cartesian, but could be, e.g., polar coordinates,
    or a transformation.
\scriptsize

```{r, fig.width=4, fig.height=2}
ggplot(gapminder,aes(x=gdpPercap,y=lifeExp,color=continent)) + 
  geom_point(alpha=0.5) + coord_trans(x="log10")
```


## Components of a `ggplot`: faceting

\small

- How to break up the data into subsets and arrange
multiple plots on the graphics device.

\scriptsize

```{r, fig.width=4, fig.height=2}
p2 +  facet_grid(continent ~ .)
```

## Why so many components?

\small

- A framework for the components of a graph.
- Gives the user the ability to change indvidual 
components one at a time.

## More details

- Layers 
    - data, aesthetic mapping, geom, statistical transformation 
    and position adjustment (to be defined)
- Tools for working with layers
- Scales, axes and legends
- Positioning: faceting and coordinate systems

## Example dataset: Diamonds

\small

- Price and quality of about 54,000 diamonds.
- Quality measures are carat (size), cut, colour and clarity
- Also included are three measures of the dimensions of 
each diamond labelled x, y and z.

\scriptsize

```{r}
data(diamonds)
head(diamonds)
```

## Initialization 

\small

- We first initialize the plot.
- Initializing is done with `ggplot()`.
    - We usually specify the default data and aesthetic 
    mappings for all subsequent layers, though this is not
    necessary.
    - Without layers, the plot is not displayed.

\scriptsize

```{r}
p <- ggplot(diamonds,aes(x=carat,y=price,colour=cut))
```

## Adding layers

\small

- Add with a `+`
- The `layer()` function can be used to specify a 
geom, stat and position 
    - data and mapping will be inherited from initialization

\scriptsize

```{r,fig.width=4,fig.height=2.5}
p + layer(geom="point", stat="identity", position="identity")
```

- The data and aesthetic mapping for the layer are
inherited from the initialization.


## Shortcuts for adding layers

\small

- Shortcut functions are of the form `geom_XXX()` and
`stat_XXX()`.
    - The `geom_XXX()` functions have a default stat and position
    - The `stat_XXX()` functions have a default geom and position
    - The `geom_XXX()` can over-ride the default stat and 
    the `stat_XXX()` can over-ride the default geom though
- Call on the previous slide is equivalent to 

\scriptsize

```{r}
p <- p + geom_point() 
```

## Aside: Plot objects

\small

- Notice that plot objects can be stored as R objects:

\tiny

```{r}
summary(p)
```



