---
title: "Stat 341 Lecture 1"
author: "Brad McNeney"
date: '2017-01-04'
output: 
  beamer_presentation:
    toc: true
---
# Introducing R

## What is R?

- R is an open-source environment for statistical computing and graphics.
    - An implementation of the S language
    (<https://en.wikipedia.org/wiki/S_(programming_language)>)
- Started in the mid-1990's by Ross Ihaka and Robert Gentlemen at Auckland University
- Now maintained by a team of experts called the R Development Core Team
- A "packages" system allows any user to bundle R code, data and examples together.
- R and R packages are distributed through the Comprehensive R Archive Network (CRAN). 
- SFU has a CRAN mirror at <http://cran.stat.sfu.ca>

## What does "environment" mean?
- R is a fully-functioning programming environment with all the usual constructs, such as 
    - conditionals (if-then-else), 
    - loops 
    - user-defined functions.
- In addition there are built-in facilities for 
    - data input, storage, manipulation, and output
    - optimization, matrix computation, etc., 
    - random number generation,
    - data analysis and graphics.
- "Base" R is good, but it is the package system that makes R great.

## R packages
- The R package system is the key to R's success
- It has allowed statisticians and other data scientists to implement and distribute their work to be used by others.
- The R package system enforces some rules about how packages are structured, but differences in programming styles of package authors mean different interfaces
    - Users need to be aware of different data structures for input, output and of different styles of graphics
    
# Course objectives
## Course objectives
- Understand basic R data structures and programming
- Learn how to use base R and R package functions for data management, exploration, presentation and analysis
- Or, ...
    - learn enough R to get by in SFU Stat courses and write "proficient in R" on your resume

# Getting started
## Getting started with R, RStudio and git
- A brief set of instructions are on the canvas page 
<https://canvas.sfu.ca/courses/30477/pages/getting-started-with-r-rstudio-and-git>
- More details can be found in Appendix B.1 of the text.


## R reference cards
Here are two. A google search will turn up many more.

- <http://cran.stat.sfu.ca/doc/contrib/Baggott-refcard-v2.pdf>
- <http://cran.stat.sfu.ca/doc/contrib/refcard.pdf>

## Starting R
- Start R by starting RStudio. 
- The "Console" window is where you can type your commands.
- However, it is good practice to open an R script, type your
commands in the script, and then submit the commands to the R console.
    - `Session -> Set Working Directory` to set the working directory
    - `File -> New File -> R Script` to open a new R script
    - type your commands into the script
    - put your cursor on the line you want to submit and hit `Ctrl-enter`
- Save your script for later use.
- More on the RStudio interface at <https://support.rstudio.com/hc/en-us/sections/200107586-Using-RStudio>

# Quick-start R 

## R as a calculator
\small
```{r}
2+4
5*6
x<-6
x*5
```
- The assignment operator `<-` assigns the value 6 to the variable x

## R for matrix computation
\small
```{r}
A <- matrix(c(1,2,3,4),nrow=2,ncol=2) 
A
x <- c(5,6)
b<- A %*% x
b
```
- Type `help(matrix)` or `?matrix` for help on the `matrix()` function.
- The `%*%` operator is matrix multiplication: `help("%*%")`.

## Matrix computation, continued
\small
```{r}
solve(A,b) # Solve Ax = b for x
solve(A) # A inverse
```

## Random number generation
\footnotesize
```{r}
set.seed(123) # optional, but makes computation reproducible
n<-100
y<-rnorm(n,sd=1)
hist(y)
```

## R data 

\footnotesize
```{r}
data(cars) # load a built-in example dataset, help(cars) for info
head(cars) # look at the first few rows
summary(cars) # summarize the variables
```

## Plotting data
\small
```{r}
plot(cars)
```

## Modelling
\footnotesize
```{r}
fit <- lm(dist~speed,data=cars) # fit a linear model, help(lm)
summary(fit)
```

## Modelling, continued
\small
```{r}
plot(cars)
abline(fit)
```

## Extracting model components 
\footnotesize
```{r}
head(residuals(fit))
vcov(fit)
```
- See the help page for more components

## Model diagnostics
```{r}
par(mfrow=c(2,2))
plot(fit)
```

# Reading
## Reading 
- Text, Appendix B
- R and RStudio help pages/reference cards mentioned in this lecture

