---
title: "Stat 341 Lecture 3"
author: "Brad McNeney"
date: '2018-01-18'
fontsize: 10pt
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```

# Subsetting with `dplyr`

## Subsetting with `dplyr`

* Subsetting of data frames in the tidyverse is done with 
the `filter()` and `select()` functions from `dplyr`.
* We will discuss filtering and selecting in much more
detail later in the course. 

\small

```{r}
library(gapminder) # contains the data
library(dplyr)
```

## 

* Take a look at the top and bottom few
lines of raw data.

\small

```{r}
head(gapminder)
```

##

\small

```{r}
tail(gapminder)
```

## 

\small

```{r}
summary(gapminder)
```

## Tibbles

* Tibbles are an implementation of data frames that is
developing as the standard data structure in the 
tidyverse.
* See the documentation for the differences between
data frames and tibbles:
https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html

* Coerce a data frame to a tibble with `as_tibble()` and 
a tibble to a data frame with `as.data.frame()`.

## Select rows with `filter()`

* Extract the data from 2007:

\small

```{r}
gapminder07 <- filter(gapminder, year == 2007)
head(gapminder07)
```

## Select columns with `select()`

* Extract the year, life expectancy and country with 
select.

\small

```{r}
gmReduced <- select(gapminder, year, lifeExp, country)
gmReduced <- select(gapminder, -continent, -pop, -gdpPercap)
```


# Control Flow

## `if` and `if`-`else` 

\small

- `if` tests a condition and executes code if the condition is true. Optionaly, can couple with an `else` to specify code to execute
when condition is false.

```{r}
if("cat" == "dog") {
  print("cat is dog")
} else {
  print("cat is not dog")
}
```


## `for` loops

\small

Example:

```{r}
n <- 10; nreps <- 100; x <- vector(mode="numeric",length=nreps)
for(i in 1:nreps) {
  # Code you want to repeat nreps times
  x[i] <- mean(rnorm(n))
}
summary(x)
print(i)
```

## `for` loop index set

\small

- Index sets of the form `1:n` are most common, but 
can be almost any atomic vector.

```{r}
ind <- c("cat","dog","mouse")
for(i in ind) {
  print(paste("There is a",i,"in my house"))
}
```

## `while` loops

\small

- Use a `while` loop when you want to continue until
some logical condition is met.

```{r}
set.seed(1)
# Number of coin tosses until first success (geometric distn)
p <- 0.1; counter <- 0; success <- FALSE
while(!success)  {
  success <- as.logical(rbinom(n=1,size=1,prob=p))
  counter <- counter + 1
}
counter
```

## `break`

- `break` can be used to break out of a `for` or `while` loop.

```{r}
for(i in 1:100) {
  if(i>3) break
  print(i)
}
```

# Reading from and writing to files

## Native format

\small

- Use `save()` to save R objects to an "R Data" file.
    - `save.image()` is short-hand to save all objects 
    in the workspace

```{r}
x <- rnorm(100); y <- list(a=1,x=x)
save(x,y,file="test.RData") # Or .rda, or ...
```

- Load R Data files into the workspace with `load()`.
```{r}
load("test.RData")
file.remove("test.RData") 
```

## Table format files

- `read.table()` is the main function for reading 
tabular data from plain-text files.
    - `read.csv()` and `read.delim()` are basically 
    `read.table()` with defaults for reading comma- and tab-
    delimited files.
- `write.table()`, `write.csv()` and `write.delim()` are
the analogous functions for writing tabular data
```{r}
write.table(matrix(1:9,3,3),file="test.txt")
test <- read.table("test.txt")
file.remove("test.txt")
test
```

## Reading files from a URL

- `load()`, `read.table()`, etc. can read data from a URL.

\scriptsize

```{r}
baseURL <- "http://people.stat.sfu.ca/~mcneney/Teaching/Stat341/"
rdURL <- url(paste0(baseURL,"Data/PorschePrice.rda"))
load(rdURL)
head(PorschePrice)
csvURL <- url(paste0(baseURL,"Data/PorschePrice.csv"))
PorschePrice <- read.csv(csvURL)
```

## Reading more complex text files

\footnotesize

- Defaults for `read.table()` are not always what you want.
    - In particular, the default for reading columns that
    include text is to coerce to a factor.
    - Also replaces spaces in column headers with `.`.

```{r}
exURL <- url(paste0(baseURL,"Data/Ex1_1_4.txt"))
ex <- read.table(exURL,header=TRUE,sep="\t") 
# same as  ex <- read.delim(exURL)
ex
str(ex)
```

## `stringsAsFactors`

\small

- Reading columns that include characters
in as factors is controlled by a global option in your R session
called `stringsAsFactors`, set to `TRUE` by default.
- If you want to set to `FALSE` for an R session type
`options(stringsAsFactors = FALSE)` into the Console.
- An alternative is to over-ride the default in the
call to `read.table()`:
```{r}
exURL <- url(paste0(baseURL,"Data/Ex1_1_4.txt"))
ex2 <- read.table(exURL,header=TRUE,sep="\t",
                  stringsAsFactors=FALSE) 
str(ex2)
```


## Post-processing: dates

- `Date.of.purchase` should be coerced to a `Date` object.

\small

```{r}
ex2$Date.of.purchase <- 
  as.Date(ex2$Date.of.purchase,"%m/%d/%Y") 
str(ex2)
diff(ex2$Date.of.purchase)
```

## Post-processing: strings

\small

- Will probably want to remove the `$` in `amount` and
coerce to numeric.
- Many options for manipulating strings -- will discuss 
in detail later in the course. 
- For now, just mention `substr()` and `paste()`: 
```{r}
maxStringLen <- 10 # allows for amounts up to $999999.99
ex2$amount <- as.numeric(
  substr(ex2$amount,start=2,stop=maxStringLen)
  )
str(ex2)
paste("$",ex2$amount,sep="") # or paste0("$",ex2$amount)
```

## Reading and writing Excel files

- If you have a working copy of Excel, you can export to
`.csv` format 
and use `read.csv()` and `write.csv()`.
- However, there are functions in several 
R packages for reading directly from an Excel file.
    - E.G., see the `read_excel()` function from the
    `readxl` package.
    - Or, try `read.xls()` from the 
    `gdata` package.

## Reading data with `readr`

* The `readr` package provides `tidyverse` equivalents to the
`read.` functions from base R.
    * The equivalents are of basically the same name, with
    the dot replaced by an underscore; e.g., 
    `read_table()` replaces `read.table()`.
* These equivalents (i) return a `tibble` and try 
to correctly parse strings and dates.

* See the documentation at http://readr.tidyverse.org/ for
full details.

##

\small

```{r}
library(readr) # or library(tidyverse)
exURL <- url(paste0(baseURL,"Data/Ex1_1_4.txt"))
ex2 <- read_tsv(exURL,col_names=TRUE) # tab-separated
ex2 # Date of purchase not coerced to Date
```


