---
title: "Stat 341 Lecture 2"
author: "Brad McNeney"
date: '2017-01-04'
fontsize: 10pt
output:  
 beamer_presentation:
  toc: true
---
# R objects
## R objects
- In R, data structures and functions are all referred to 
as "objects".
- Objects are created with the assignment operator `<-`; e.g.,
`x <- 1`.
    - The objects a user creates from the R console are contained in
    the user's workspace, called the global environment. 
    - Use `ls()` to see a list of all objects in the workspace. 
- We will discuss data structures first, and then functions and
R packages.
- In between we will discuss logical and relational operators that 
can be used to subset data structures.

# R Data Structures
## R Data Structures

- Reference Wickham (2014), Advanced R, Chapter 2
- Focus on four common data structures: atomic vectors, lists, matrices and data frames.
- Atomic vectors and lists are 1d, while matrices and data frames are 2d objects
- R has no true scalars; e.g., in `x<-1`, `x` is a vector of length one.
- R also has an array data structure for higher dimensional elements that we will not discuss.
- Use `str()` to see the structure of an object

# Vectors 

## Vectors
\small

- Vectors can be either atomic or list
    - atomic vectors must be comprised entirely of logical, integer, double (numeric) or character elements
    - lists can be comprised of multiple data types
- Empty vectors can be created by the `vector()` function:
```{r}
avec <- vector(mode="numeric",length=4)
lvec <- vector(mode="list",length=4)
```
- Data vectors can be created with `c()` or `list()`:
```{r}
avec <- c(50,200,77)
lvec <- list(50,200,77,c("grey","thin"))
```

## Combining vectors
\small

- Use `c()` to combine vectors
```{r}
c(avec,c(100,101))
c(lvec,TRUE)
```

## Vector attributes
\small

- Vectors have a type and length and, optionally, attributes such as names. 
```{r}
names(lvec) = c("age","weight","height","hair")
str(lvec)
```
- We can specify element names when creating a vector; e.g.:
```{r}
lvec <- list(age=50,weight=200,height=77,hair=c("grey","thin"))
```

## Factors
\small

- The statistical concept of a factor is important in experimental 
design. 
- Factors are implemented in R as atomic vectors with attributes
`class` and `levels`:
```{r}
trt <- factor(c("drug1","placebo","placebo","drug2"))
attributes(trt)
str(trt)
```
- The levels are coded numerically (1, 2 and 3)
with assigned labels ordered alphabetically
("drug1", "drug2" and "placebo")

## More on object class
\small

- You can create your own `class` for an object. 
- Such "meta-data" can be used to tell R how to handle the 
object; e.g., how to print it, summarize it, etc.
```{r}
class(lvec) <- "prof"
lvec
```


## Subsetting vectors and extracting elements
\small

- Subset with `[` or by name:
```{r}
lvec[c(1,3)] # same as lvec[c("age","height")]
```
- Extract individual elements with `[[`, or `$` for named objects:
```{r}
lvec[[4]]
lvec$hair
```

## Subsetting factors
\small

- Subsetting may remove all instances of a level, but the 
level will be retained in the data structure
```{r}
trt[1:3]
```
- If subsetting is intended to remove a level of the factor, use `drop=TRUE`
```{r}
trt[1:3,drop=TRUE]
```

## Subsetting and assignment
\small

- You can combine subsetting and assignment to change the value of 
vectors
```{r}
avec
avec[2] <- 210
avec
```

## Coercion: atomic vectors to lists
\small

- Atomic vectors can be coerced to lists with `as.list()`:
```{r}
avec = c(age=50,weight=200,height=77)
avec
as.list(avec)
```

## Coercion: lists to atomic vectors
\small

- Lists can be "flattened" into atomic vectors with `unlist()`:
```{r}
unlist(lvec)
```
- Notice how the numeric values are coerced to the more flexible 
character type. 
- The order of flexibility, from least to most, is logical,
integer, numeric, character.

## Coercion: factors to atomic vectors
\small

- We saw how to use `factor()` to coerce an atomic vector to a factor.
- Use `as.vector()` to coerce a factor back to an atomic vector. 
- The result is a character vector. You may need to use `as.numeric()` to coerce to numeric, if required.
```{r}
a <-factor(c(2,1,1,2))
as.vector(a)
as.numeric(as.vector(a))
```


# Matrices and data frames
## Matrices and data frames
- Though both 2d objects, matrices and data frames
are different enough that we will need to discuss them 
separately.
- The elements of a matrix must all be of the same type.
- Data frames are essentially lists where each list element
has the same length. Thus data frames can include columns
of varying type.

## Matrices
\footnotesize

- Matrices can be created with the `matrix()` function as in 
```{r}
A <- matrix(1:4,nrow=2,ncol=2) 
A
```
- Here `1:4` is the same as `c(1,2,3,4)`
- The default is to read the data vector into the matrix column-by-column. To 
read row-by-row instead use the `byrow=TRUE` argument:
```{r}
A <- matrix(1:4,nrow=2,ncol=2,byrow=TRUE) 
A
```

## Combining matrices
\small

- Combine matrices with `rbind()` and `cbind()`:
```{r}
rbind(A,matrix(c(5,6),nrow=1,ncol=2))
cbind(A,A)
```

## Matrix attributes
\footnotesize

- Matrices have a type, dimension (number of rows, number of 
columns) and optional attributes such as dimnames (row and column names).
```{r}
dim(A)
nrow(A)
colnames(A) <- c("var1","var2")
rownames(A) <- c("subj1","subj2")
A
str(A)
```

## Subsetting matrices 
\footnotesize

- Subset with `[` and a comma to separate rows from columns:
```{r}
A[1,1]
A[1,]
A[,1]
```
- When a subsetting operation leads to a vector, the dimension 
of the object is "dropped" from 2 to 1. To prevent this use
`drop=FALSE`:
```{r}
A[1,,drop=FALSE]
```


## Extracting elements from matrices 
\footnotesize

- Can use `[[` to extract elements, but this is not necessary 
because of the way subsetting to a single element drops to 
a vector of length 1 by default:
```{r}
A[[1,1]]
A[1,1]
```

## Coercion: Matrices to/from vectors
- We have already seen how `matrix()` coerces a vector to a matrix
- `as.vector()` applied to a matrix creates a vector by concatenating
columns:
```{r}
as.vector(A)
```

##  Data frames
\small

- Data frames (class `data.frame`) are the usual way to 
store data in R.
    - Rows are intended to be observational units, columns variables
    - Implemented as a list (columns are list elements), but also behave like a matrix in terms of combining and subsetting.
- Create with `data.frame`:
```{r}
set.seed(1)
n <- 4
x <- 1:n; y <- rnorm(n,mean=x,sd=1) # multiple commands separated by ;
dd <- data.frame(x=x,y=y) # like making a list
str(dd)
```

## Subsetting and combining data frames
\small

- Can subset columns like a list:
```{r}
dd$x # or dd[[1]]
```

- Can subset columns/rows and combine like matrices; e.g.,
```{r}
dd[1:2,]
zz = data.frame(z=runif(4))
cbind(dd,zz)
```


# Logical  and relational operators
## Logical operators
\small

- The basic logical operators are described in `help("Logic")`.
- `!` is NOT
- `&` and `&&` are AND, with `&` acting vector-wise and `&&` acting 
on scalars
- `|` and `||` are OR, with `|` acting vector-wise and `||` acting 
on scalars
- Make sure you understand the following:
```{r}
x <- c(TRUE,TRUE,FALSE); y <- c(FALSE,TRUE,TRUE)
!x ; x&y ; x&&y ; x|y ; x||y
```

## Relational operators
\small

- Relational operators can be used to compare values in atomic vectors
    - See `help("Comparison")`
- `>` is greater than, `>=` is greater than or equal
- `<` is less than, `<=` is less than or equal
- `==` is equal and `!=` is not equal
- Make sure you understand the following:
```{r}
x <- 1:3; y <- 3:1
x>y ; x>=y ; x<y ; x<=y ; x==y ; x!=y
```


## Subsetting vectors with logical expressions
\small

- Can subset with logicals and `[`:
```{r}
avec
avec>100
avec[avec>100]
avec[avec>50 & avec<100]
```

## Subsetting matrices with logical expressions

- Can also subset matrices, but results may not be as expected:
```{r}
A
A>1
A[A>1] # coerces to a vector
```

## Subset and assign with logical expressions
\footnotesize

- Combine subset and assign to change the value of objects
```{r}
A[A>1] <- 9
A
```
- In the above substitution, the vector 9 is shorter than the 
three elements in `A>1` so R "recycles" the 9 three times. 
- Be careful about recycling:
```{r}
A[A>1] <- c(-10,10) # Throws a warning 
A # R used c(-10,10), then just the -10 
```

# Aside: Special values

## Missing values
\small

- R has a special data code for missing data: `NA`
- Test for and set missing values with `is.na()`
```{r}
avec
is.na(avec)
is.na(avec) <- 2 
avec
```



## Infinite and undefined values
\small

- R has a special codes for infinite 
values (`Inf`) and undefined values (`NaN`).

- Test for `Inf` and `NaN` with `is.infinite()` and `is.nan()`.
```{r}
ii <- 1/0 ; nn <- 0/0
ii
is.infinite(ii)
nn
is.nan(nn)
```

# R Functions

## R functions
\small

- Reference Wickham (2014), Advanced R, Chapter 6
- In R, functions are objects with three essential components: 
    - the code inside the function, or `body`,
    - the list of arguments to the function, or `formals`, and
    - a data structure called an `environment` which is like a 
    map to the memory locations of all objects defined
    in the function.
    
- Functions can have other attributes, but the above three are 
essential.

## Example function 
\small

```{r}
f <- function(x) {
  return(x^2)
}
f
```

## The function body
\small

- This is the code we want to execute. 
- When the end of a function is reached without a call to 
`return()`, the value of the last line is returned.
    - So in our example function, we could replace `return(x^2)` 
    with just `x^2.

## The function formals
\small

- These are the arguments to the function. 
- Function arguments can have default values, as in:
```{r}
f <- function(x=0) { x^2} 
```
- Argument defaults can be defined in terms of other arguments:
```{r}
f <- function(x=0,y=3*x) { x^2 + y^2 }
f()
f(x=1)
f(y=1)
```

## Argument matching when calling a function
\footnotesize

- When you call a function, the arguments are matched first by 
name, then by "prefix" matching and finally by position:
```{r}
f <- function(firstarg,secondarg) {
  firstarg^2 + secondarg^2
}
f(firstarg=1,secondarg=2)
f(s=2,f=1)
f(2,f=1)
f(1,2)
```

## The function environment
\small

- The environment within a function is like a map to the 
memory locations of all its variables.
- The function arguments are "passed by value", meaning that
a copy is made and stored in the function's environment.
- Variables created within the function are also store in its
environment
```{r}
f <- function(x) {
  y <- x^2
  ee <- environment() # Returns ID of environment w/in f
  print(ls(ee)) # list objects in ee
  ee
}
f(1) # function call
```

## Enclosing environments
\small

- Our function `f` was defined in the global environment,
`.GlobalEnv`,
which "encloses" the environment within `f`.
- If `f` needs a variable and can't find it whithin
`f`'s environment, it will look for it in the enclosing 
environment, and then the enclosing environment of `.GlobalEnv`,
and so on. 
- The `search()` function lists the heirarchy of 
environments that enclose `.GlobalEnv`. 
```{r}
search()
```
- To facilitate this search, each environment includes 
a pointer to its enclosing environment.

# R packages

## Loading packages
\small

- Use the `library()` command to load packages.
- When we load a package
it is inserted in position 2 of the search list, 
just above `.GlobalEnv`.
```{r}
# install.packages("hapassoc")
library(hapassoc) 
search()
```
- Detach a package from the search list with `detach()`
```{r}
detach("package:hapassoc")
search()
```

## Package namespaces
\small

- Package authors create a list of objects
that will be visible to users when the package
is loaded. This list is called  the package namespace.
- You can access functions in a package's namespace 
without loading the package using
the `::` operator.
```{r}
set.seed(321)
n<-30; x<-(1:n)/n; y<-rnorm(n,mean=x); ff<-lm(y~x)
car::sigmaHat(ff)
```
- Doing so does not add 
the package to the search list.