---
title: "Stat 341 Lecture 3"
author: "Brad McNeney"
date: '2017-01-18'
fontsize: 10pt
output:  
 beamer_presentation:
  toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Control Flow

## `if` and `if`-`else` 

\small

- `if` tests a condition and executes code if the condition is true. Optionaly, can couple with an `else` to specify code to execute
when condition is false.

```{r}
if("cat" == "dog") {
  print("cat is dog")
} else {
  print("cat is not dog")
}
```


## `for` loops

\small

Example:

```{r}
n <- 10; nreps <- 100; x <- vector(mode="numeric",length=nreps)
for(i in 1:nreps) {
  # Code you want to repeat nreps times
  x[i] <- mean(rnorm(n))
}
summary(x)
print(i)
```

## `for` loop index set

\small

- Index sets of the form `1:n` are most common, but 
can be almost any atomic vector.

```{r}
ind <- c("cat","dog","mouse")
for(i in ind) {
  print(paste("There is a",i,"in my house"))
}
```

## `while` loops

\small

- Use a `while` loop when you want to continue until
some logical condition is met.

```{r}
set.seed(1)
# Number of coin tosses until first success (geometric distn)
p <- 0.1; counter <- 0; success <- FALSE
while(!success)  {
  success <- as.logical(rbinom(n=1,size=1,prob=p))
  counter <- counter + 1
}
counter
```

## `break`

- `break` can be used to break out of a `for` or `while` loop.

```{r}
for(i in 1:100) {
  if(i>3) break
  print(i)
}
```

# Reading from and writing to files

## Native format

\small

- Use `save()` to save R objects to an "R Data" file.
    - `save.image()` is short-hand to save all objects 
    in the workspace

```{r}
x <- rnorm(100); y <- list(a=1,x=x)
save(x,y,file="test.RData") # Or .rda, or ...
```

- Load R Data files into the workspace with `load()`.
```{r}
load("test.RData")
file.remove("test.RData") 
```

## Table format files

- `read.table()` is the main function for reading 
tabular data from plain-text files.
    - `read.csv()` and `read.delim()` are basically 
    `read.table()` with defaults for reading comma- and tab-
    delimited files.
- `write.table()`, `write.csv()` and `write.delim()` are
the analogous functions for writing tabular data
```{r}
write.table(matrix(1:9,3,3),file="test.txt")
test <- read.table("test.txt")
file.remove("test.txt")
test
```

## Reading files from a URL

- `load()`, `read.table()`, etc. can read data from a URL.
\scriptsize

```{r}
baseURL <- "http://people.stat.sfu.ca/~mcneney/Teaching/Stat341/"
rdURL <- url(paste0(baseURL,"Data/PorschePrice.rda"))
load(rdURL)
head(PorschePrice)
csvURL <- url(paste0(baseURL,"Data/PorschePrice.csv"))
PorschePrice <- read.csv(csvURL)
```

## Reading more complex text files

\footnotesize

- Defaults for `read.table()` are not always what you want.
    - In particular, the default for reading columns that
    include text is to coerce to a factor.
    - Also replaces spaces in column headers with `.`.

```{r}
exURL <- url(paste0(baseURL,"Data/Ex1_1_4.txt"))
ex <- read.table(exURL,header=TRUE,sep="\t") 
# same as  ex <- read.delim(exURL)
ex
str(ex)
```

## `stringsAsFactors`

\small

- Reading columns that include characters
in as factors is controlled by a global option in your R session
called `stringsAsFactors`, set to `TRUE` by default.
- If you want to set to `FALSE` for an R session type
`options(stringsAsFactors = FALSE)` into the Console.
- An alternative is to over-ride the default in the
call to `read.table()`:
```{r}
exURL <- url(paste0(baseURL,"Data/Ex1_1_4.txt"))
ex2 <- read.table(exURL,header=TRUE,sep="\t",
                  stringsAsFactors=FALSE) 
str(ex2)
```


## Post-processing, part I

- `Date.of.purchase` should be coerced to a `Date` object.
```{r}
ex2$Date.of.purchase <- 
  as.Date(ex2$Date.of.purchase,"%m/%d/%Y") 
str(ex2)
diff(ex2$Date.of.purchase)
```

## Post-processing, part II

\small

- Will probably want to remove the `$` in `amount` and
coerce to numeric.
- Many options for manipulating strings. 
- Useful functions, in increasing
order of flexibility and complexity of use are
    - `substr()`
    - `strsplit()`
    - `gsub()`
- Will illustrate the simplest: 
```{r}
maxStringLen <- 6 # allows for amounts up to $99.99 only
ex2$amount <- as.numeric(substr(ex2$amount,2,maxStringLen))
str(ex2)
```
- See `help(substr)` for a description.
