---
title: "Lecture 10"
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```

## Load packages

\small

```{r}
library(dplyr)
library(broom)
library(Stat2Data)
data(AutoPollution)
data(CloudSeeding)
data(Hawks)
```

# Random number generation

## Random number generation

- Reference: Chapter 2 of Robert and Casella (2004),
Introducing Monte Carlo Methods with R
    - Online copy available from the library: [http://search.lib.sfu.ca/?q=introducing%20monte%20carlo%20methods%20with%20r]

## Why generate random numbers?

\small

- Simulating from a data-generating process allows us to evaluate
the performance of methods for making inference from such data.
    - E.G., to evaluate the performance of least squares regression 
    for inference of regression coefficients, we can repeatedly 
    generate datasets, perform inference, and summarize the results
    (example later).
- Many modern statistical methods are simulation-based; e.g.,
    - permutation tests
    - boostrap tests and confidence intervals
    - cross validation for selecting "tuning" parameters 
    
## Random number generation

\small

- We can't generate truly random numbers. 
- Instead we generate deterministic sequences or *streams* of 
pseudo-random numbers.
    - Goal: Every $n$-tuple from the stream should be
    statistically indistinguishable from a random sequence
    of size $n$ drawn from a uniform distribution on (0,1), denoted
    $U(0,1)$.
- Basic approach is to
    - Generate $U(0,1)$ deviates
    - Transform to a sample from the distribution of interest.
- Generating $U(0,1)$ deviates
    - Several random number generators (RNGs) have been developed
    - No consensus on which is "best"
    - We'll use the default in R

## Random number seed

\small

- The current position in the 
random number stream is called the *seed*.
- Information on the seed is contained in the variable `.Random.seed` 
in your workspace.
    - If we don't set the seed, R will do it for us.
    - Each time we generate random numbers, the seed is incremented.
    - When we save the workspace, the seed is saved too.
    - **Warning:** If you start R from a previously-saved workspace,
    you will get the random seed. Starting multiple simulations 
    from the same seed gives the same simulations
- You can set the seed with `set.seed()`.

## Setting the seed

\scriptsize

```{r}
set.seed(123)
runif(10)
set.seed(123)
runif(10)
set.seed(42)
runif(10)
```

## Generating uniforms

\scriptsize

```{r, fig.width=4,fig.height=3}
Nsim <- 1000
x <- runif(Nsim)
hist(x)
```


## Example: Assessing uniformity with a test

```{r}
ks.test(x,"punif")
```

## Generating from other distributions

\small

- For common distributions, we don't have to work out the transformations 
to convert from $U(0,1)$ deviates to the distribution of interest.
- Built-in functions in R have names like `rnorm()`, `rt()`, 
`rchisq()`

## Example: Generating from a Normal distribution

\small

```{r}
Nsim <- 1000; mu <- 0; sig <- 1
x <- rnorm(Nsim,mean=mu,sd=sig)
```

## Example: Assessing normality graphically

\scriptsize

```{r, fig.width=4,fig.height=3}
qqnorm(x); abline(a=0,b=1,col="red")
```

## Example: Assessing normality with a test

```{r}
ks.test(x,"pnorm")
```

## Application: Demonstrate the central limit theorem (CLT)

\small

- The CLT says that the sampling distribution of an average is
approximately normal for large sample sizes, regardless of 
the parent population from which the sample is drawn.
- Example data-generating process: Samples of 
size $n=30$ from an exponential distribution with rate 5.

\scriptsize

```{r,fig.width=4,fig.height=2.7}
x <- rexp(n=30,rate=5)
hist(x)
```

## CLT: Sampling distribution of averages

\scriptsize

```{r,fig.width=4,fig.height=2.5}
library(ggplot2)
set.seed(8675309)
xbar <- vector(mode="numeric",length=Nsim)
for(i in 1:Nsim) {
  xbar[i] <- mean(rexp(n=30,rate=5))
  }
ggplot(data.frame(xbar=xbar),aes(sample=xbar)) + geom_qq() + geom_qq_line()
```


## `replicate()`

\small

- The `replicate()` function can be used for simulation.
- The following is equivalent to the `for` loop in the 
previous slide.

\scriptsize

```{r}
set.seed(8675309)
expMean <- function(n=30,rate=5) { mean(rexp(n=n,rate=rate)) }
xbar2 <- replicate(Nsim,expMean())
all.equal(xbar,xbar2)
```

## Application: Simulation study

\small

- We can conduct a simulation study to evaluate the 
type 1 error of the t test of the null hypothesis that
a regression coefficient is zero.
    - Simulate under the null hypothesis, but with non-normal
    errors
- Fix 
    - sample size `n=30`
    - values of a single covariate `x=(1:n)/n`
- Replicate the following simulation `Nsim` times
    - Simulate responses $Y$ as $Z-1$ where 
    $Z$ has as exponential distribution with rate 1
    (mean 1).
    - Fit the regression model by least squares and 
    return the p-value from the test of association with `x`.
- Calculate the proportion of p-values less than 0.05.
    - Should be 5\%
    
## Simulation code: simulating one dataset

\scriptsize

```{r}
n <- 30; x <- (1:n)/n
y <- rexp(n)-1
fit <- lm(y~x)
tidy(fit)
tidy(fit) %>% filter(term=="x") %>% select(p.value) 
# or 
tidy(fit)[2,5]
```

## Simulation study

\scriptsize

```{r}
Nsim <- 1000
simfunc <- function() {
  y <- rexp(n)-1
  tidy(lm(y~x))[2,5]
}
simout <- replicate(Nsim,simfunc())
alphaNominal <- 0.05
alphahat <- mean(simout<alphaNominal)
alphahat
SE <- sqrt(alphahat*(1-alphahat)/Nsim)
c(alphahat - 1.96*SE, alphahat + 1.96*SE)
```

## Note on simulation error

- Use the normal approximation 
to the mean of `Nsim` binary variables.
    - Let $\alpha$ be the true type 1 error of the procedure,
    and $\hat{\alpha}$ be our simulation-based estimate
    - $\hat{\alpha}$ is a mean and can be shown to 
    be approximately Normal with mean $\alpha$
    and SD $\sqrt{\alpha(1-\alpha)/Nsim}$.
    - SE is obtained by plugging in $\hat{\alpha}$
- Can define simulation error as the margin of 
error in a 95\% confidence interval for the 
parameter being estimated.
    - In our study, the CI covers 0.05, so we say
    that the empirical type 1 error is within
    simulation error of the nominal level.


## Sampling with and without replacement

\small

- Draw probability weighted samples of size $n$ from a set with `sample()`
    - Sampling can be with replacement or without.

\scriptsize

```{r}
myset <- 1:10; nset <- length(myset)
probwts <- rep(1/nset,nset) 
n<-8
sample(myset,size = n, replace = TRUE, prob=probwts)
sample(myset,size = n, replace = FALSE, prob=probwts)
```

## Notes on sampling with and without replacement

\small

- Weights need not sum to one
    - They will be normalized
- If sampling with replacement, the size of the 
sample can't exceed the size of the set.
- Sampling $n$ without replacement from a set of size $n$ 
is a *permutation* of the set.
    - Using all the defaults of `sample()` gives a 
    permutation.

\scriptsize

```{r}
sample(1:10)
sample(c("cat","dog","fish","zebra"))
```

# Permutation tests

## Example

\small

- Makes most sense in the context of a designed experiment.
- Example experiment: 
    - 36 cars were randomly assigned to receive one
    of two types of car muffler. 
    - The noise in decibels of each car was measured. 
    - Is there a difference in sound level between the 
    two mufflers?
- A standard ANOVA for these data is as follows.

```{r}
mfit <- lm(Noise ~ Type, data=AutoPollution)
tidy(anova(mfit)) # F stat is 1,5 element
Fstat <- tidy(anova(mfit))[1,5]
```

## Permutation distribution of F

\small

- Under the null hypothesis the noise of cars 
doesn't depend on the muffler type, we just have
36 cars assigned randomly to two groups.
    - Cars are said to be *exchangeable* (with repect to noise)
    - The distribution of the F statistic under
    all possible re-randomizations is the 
    randomization distribution.
- Randomization can be achieved by permuting the `Noise`
variable.
    - The randomization distribution can also be called
    a permutation distribution and the test that 
    compares the observed F to this distribution a
    permutation test.

## Permutation test on example

\scriptsize

```{r}
fstatPerm <- function() {
  permdat <- data.frame(Type = AutoPollution$Type, 
                        Noise = sample(AutoPollution$Noise))
  mm <- lm(Noise ~ Type, data=permdat)
  tidy(anova(mm))[1,5]
}
set.seed(1)
pdist <- replicate(1000,fstatPerm())
mean(pdist > Fstat) # permutation p-value
```

## Exchangeability is key assumption

\small

- Responses must be exchangeable under the null hypothesis
of no treatment effect
- Cloud seeding data
    - Clouds randomly seeded (S) or not (U)
    - Clouds from several areas of Tasmania; focus on 
    Tasmania East (TE).

\scriptsize

```{r}
CloudSeeding %>% group_by(Seeded) %>% 
  summarize(n=n(),mean=mean(TE),sd=sd(TE))
```

- Unseeded clouds are more variable in rainfall, so
they are not exchangeable.

