---
title: "Leture 12"
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

## Preparation for the final exam

* Final exam is from 8:30 - 11:30, Tuesday April 16,
in Images Theatre.

* Though the time slot for the final is three hours, I 
will aim for a final that will take most people two hours.
    * A bit less than twice as long as the midterm.

* The exam is cumulative, but with about 2/3 emphasis on 
material after the midterm (lectures 7-11) and 1/3 from before
(lectures 1-6).

* In cases where we discussed both base-R and tidyverse approaches
to a task, you are responsible only for the tidyverse version.

* The exam is closed book. R cheatsheets will be provided.
    * Cheatsheets for this year's exam available at
    https://canvas.sfu.ca/courses/43617/files/9846163/download?wrap=1
    
## Course objectives -- recap

- Understand basic R data structures and programming
- Learn how to use base R and R package functions for data management, exploration, presentation and analysis
- Learn how to use packages from the "tidyverse", a collection of modern tools for data science.
    * <https://www.tidyverse.org/>

## Overview of lectures

Focus on the following topics from lectures 1-11:

1. (no topics -- introduction and getting started)
2. dataframes, lists, vectors, functions
3. subsetting with $, [ and [[ and with dplyr, for loops, reading data from files
4. transforming variables, working with factors, working with dates
5. working with strings
6. reading from databases, merging/joining database tables and dataframes
7. what is tidy data, reshaping with gather and spread (homework 2), split-apply-combine for transformations and data summaries
8. iterating with map, graphics with ggplot2
9. graphics with ggplot2
10. pseudo-random number generation, permutation tests,
the replicate function for simulation
11. the bootstrap, cross-validation

## More Examples 

* using ggplot, gather, split-apply-combine, map

```{r}
library(tidyverse)
```

##

\scriptsize

```{r}
library(ISLR); data(Wage)
ggplot(Wage,aes(x=age,y=wage)) + geom_point(alpha=.1)
```

##

\scriptsize

```{r}
fit <- lm(wage ~ poly(age,4),data=Wage,model=TRUE)
summary(fit)$coef
plotfit <- function(fit,dat,newdat){
  preds <- data.frame(newdat,
          predict(fit,newdata=newdat,interval="confidence"))
  ggplot(dat,aes(x=age)) + geom_point(aes(y=wage),alpha=0.1) +
    geom_ribbon(aes(ymin=lwr,ymax=upr),
                data=preds,fill="blue",alpha=.2) + 
    geom_line(aes(y=fit),data=preds,color="blue")
}
```

##

\scriptsize

```{r}
age <- sort(unique(Wage$age)) 
newdat <- data.frame(age=age)
plotfit(fit,Wage,newdat)
```


##

\scriptsize

```{r}
age <- sort(unique(Wage$age))
Xmat <- data.frame(age=age,poly(age,4))
Xlong <- gather(Xmat,basisFun,b,-age) 
ggplot(Xlong,aes(x=age,y=b,color=basisFun)) + geom_line()
```

##

\scriptsize

```{r}
Wage %>% group_by(maritl) %>% summarize(n=n())
Wage <- mutate(Wage,maritl2 = fct_lump(maritl,n=2))
Wage %>% group_by(maritl2) %>% summarize(n=n())
```

##

\scriptsize

```{r}
Wage %>% split(.$maritl2) %>%
  map(~lm(wage~poly(age,4),data=.)) 
```

##

\scriptsize

```{r}
Wage %>% split(.$maritl2) %>%
  map(~lm(wage~poly(age,4),data=.)) %>%
  map_dbl(~ mean(.$residuals^2))
```

##

\scriptsize

```{r}
data(iris)
iris %>% group_by(Species) %>%
  summarize(n=n(),
            meanSL = mean(Sepal.Length),
            meanSW = mean(Sepal.Width),
            SDSL = sd(Sepal.Length),
            SDSW = sd(Sepal.Width))
```

##

\scriptsize

```{r}
set.seed(1)
iris <- iris %>% 
  group_by(Species) %>%
  sample_n(size=5) %>%
  ungroup() 
library(ggplot2)
ggplot(iris,aes(x=Sepal.Length,y=Sepal.Width,label=Species)) + geom_text()
```
