---
title: "Week 9 exercise"
author: "Brad McNeney"
date: '2018-03-12'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data on SFU Stat and Act Sci students

Questions 1, 2 and 5 include **quiz questions** that 
you must **answer on canvas** by Wednesday March 21 at
5 pm.


- The following data are from a summary of 
Stat and Act Sci students made available
by SFU's Institutional Research and Planning office
at [https://www.sfu.ca/content/dam/sfu/irp/departments/stat_tables.pdf].

```{r, warning=FALSE, message=FALSE}
stat <- data.frame(
  year = 12:16,
  FTEs = c(446.47,484.8,483.53,443.97,466),
  majors = c(213,245,260,228,233),
  minors = c(17,27,62,77,111),
  p.intl = c(.4,.42,.51,.53,.54)) 
head(stat)
```

- `year` is the end-year of an academic year; e.g., 
12 = 2011/2012, and so on up to 16 = 2015/2016.
- `FTEs` stands for full-time equivalent, which is 30 credit
hours of classes in statistics and actuarial science.
    - For example, if there are 120 students in Stat 341,
    each receiving 2 credit hours, then this amounts to 
    $120\times 2/30 = 8$ FTEs.
- `majors` is the number of declared Stat or Act Sci majors.
- `minors` is the number of declared Stat or Act Sci minors. 
- `p.intl` is the proportion of interantional students.

1. Load the packages in the `tidyverse` and 
reshape the `FTEs`, `majors` and `minors`
columns into a key-value pair of columns
with the key variable labeled as 
`type` and value variable labeled as `students`.
Which of the following commands does the reshaping?
    a) `statlong <- gather(stat,type,students,FTEs,majors,minors)`
    b) `statlong <- gather(stat,type,students,-year,-p.intl)`
    c) both of the above

2. Plot `students` versus `year` by group `type`,
with different colors for each type of student.
Add a scatterplot smoother **without** standard
error bands.
Which interpretation of the graph do you agree with?
    a) FTEs and majors are roughly steady, but minors
    are increasing.
    b) Majors are increasing, but minors are roughly steady.
    c) Both majors and minors are roughly steady.

3. Add the following variables to the `stat` data frame:

- `total = majors+minors`
- `international = total*p.intl`
- `domestic = total*(1-p.intl)`

4. Reshape the `total`, `international` and `domestic`
columns into key-value pairs with key labeled
`type` and value labeled `students`. 

5. Plot  `students` versus `year` by group `type`,
with different colors for each type of student.
Add a scatterplot smoother **without** standard
error bands.
Which is the best interpretation of the graph:
    a) The total number of majors + minors is increasing, 
    and this increase is due to a roughly equal increase
    in international and domestic students.
    b) The total number of majors + minors is increasing 
    primarily due to an increase
    in international students.