---
title: "Week 11 exercises"
author: "Brad McNeney"
date: '2018-03-29'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Obtain the data, load packages, set the random seed

We will use a dataset on fruit flies that is made available in the package
`Stat2Data`. Install `Stat2Data` and load the `FruitFlies` dataframe.
You can read about the experiment with `help("FruitFlies")

```{r}
library(Stat2Data)
data("FruitFlies")
library(ggplot2)
library(dplyr)
library(broom)
set.seed(123)
```

## Questions

1. Plot histograms of `Longevity` in separate facets for each level
of the factor `Treatment`. Use a binwidth of 10 for your histograms.

2. Do boxplots of `Longevity` by `Treatment`

3. Calculate sample size, mean and SD of `Longevity` within each 
level of `Treatment`.

4. Use `stat_summary()` to
plot 95\% confidence intervals for the mean `Longevity` in each
treatment group.

5. Estimate the permutation distribution for the $F$ test of 
treatment effect using 1000 permutations. Compute the p-value
from your permutation distribution and compare to the model-based
p-value.