---
title: "Week 10 exercises"
author: "Brad McNeney"
date: '2018-03-21'
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
library(tidyverse)
library(stringr)
library(ggplot2)
```

## Graphing youth unemployment data

1. Read the youth unemployment data in the file
`API_ILO_country_YU.csv` in the `Exercises` folder of GitHub
into a data frame called `youthUI`. Be sure to 
avoid reading strings in as factors.

2. Reshape the data on different years into key-value pairs 
with key `year` and value `unemplRate`. Remove the `X`'s from
the year names and coerce to numeric.


3. Plot unemployment rates by year 
for each "country" in `youthUI`. Represent each time series
by a line. Use an appropriate alpha level to manage overplotting.

4. Extract the subset of "Countries" whose `Country.Name`
contains the string
"(IDA & IBRD countries)", and save in a data frame
`youthDevel`. Remove the "(IDA * IBRD countries)" from 
the country names. Remove `Country.Code` from the data frame.
Notes: IDA stands for International Development Association.
Countries that qualify for IDA loans are considered among
the poorest developing countries in the world. 
IBRD stands for International Bank
for Reconstruction and Developent. IBRD countries are considered
middle-income developing countries.


5. Initialize a plot of unemployment rates by year 
for each region in `youthDevel` with different colors
and symbols for each region. Then add the following layers:
    - Add lines.
    - Add points. Do this with `geom_point()`, but add
    a comment to your R code that would generate the 
    same layer with the generic `layer()` function.
    - In the legend of your plot, modify the legend title 
    from its default to `Region`.
    - Add an informative title.
    - Add the world-wide unemployment data from `youthUI` (`Country.Name==World`).
    - Finally, display your plot.
    
